---
title: "RNA-seq TF activity computation"
author: 
  - name: "Max Trauernicht"
    email: "m.trauernicht@nki.nl"
    affiliation: "Netherlands Cancer Institute - van Steensel lab"
date: '`r format(Sys.time(), "%d/%m/%Y")`'
output: 
  html_document:
    theme: united
    highlight: pygments
    fig_caption: yes
    code_folding: hide
    df_print: kable
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
---

---

### Aim
I want to compute TF activities from RNA-seq using two strategies. 1) Differential TF expression using DESeq2. 2) VIEPER, a method to infer TF activities from gene expression data.

---


## Setup {.tabset}

<!-- little HTML script to do indentation of the table of contents -->
<script>
    $(document).ready(function() {
      $items = $('div#TOC li');
      $items.each(function(idx) {
        num_ul = $(this).parentsUntil('#TOC').length;
        $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
      });

    });
</script>

```{css, echo = FALSE}
div.sourceCode {
  overflow-x: hidden;
}
```

### Libraries

```{r set-up, message = F, warning = F}

# Load dependencies
library(tidyverse)
library(GenomicRanges)
library(rtracklayer)
library(ggplot2)
library(ggbeeswarm)
library(DESeq2)
library(RColorBrewer)
#library(GGally)
#library(ggpubr)
#library(biomaRt)
#library(clusterProfiler)
#library(pathview)
#library(enrichplot)
#library(DOSE)
library(ggnewscale)
#library(ggridges)
#library(europepmc)
library(scales)
#library(GENOVA)
library(ggh4x)
library(SummarizedExperiment)
#library(seriation)
library(dorothea)
library(decoupleR)
```

### Custom functions

```{r functions}

ExonLength <- function(genes) {
  # Get the combined exon length for the genes
  tib <- as_tibble(genes) %>%
    filter(type == "exon") %>%
    mutate(gene_id = factor(gene_id, levels = unique(gene_id))) %>%
    group_by(gene_id) %>%
    dplyr::summarise(exon_number = n(),
              exon_length = sum(width))
  
  tib
}

# From Fede:
# ggpairs custom functions
corColor <- function(data, mapping, color = I("black"), sizeRange = c(1, 3), ...) {

  x   <- eval_data_col(data, mapping$x)
  y   <- eval_data_col(data, mapping$y)
  r   <- cor(x, y)
  rt  <- format(r, digits = 3)
  tt  <- as.character(rt)
  cex <- max(sizeRange)

  # helper function to calculate a useable size
  percent_of_range <- function(percent, range) {
    percent * diff(range) + min(range, na.rm = TRUE)
  }

  # plot correlation coefficient
  p <- ggally_text(label = tt, mapping = aes(), xP = 0.5, yP = 0.5,
                   size = I(percent_of_range(cex * abs(r), sizeRange)), color = color, ...) +
    theme(panel.grid.minor=element_blank(),
          panel.grid.major=element_blank())

  corColors <- RColorBrewer::brewer.pal(n = 7, name = "RdYlBu")[2:6]

  if (r <= boundaries[1]) {
    corCol <- corColors[1]
  } else if (r <= boundaries[3]) {
    corCol <- corColors[2]
  } else if (r < boundaries[3]) {
    corCol <- corColors[3]
  } else if (r < boundaries[4]) {
    corCol <- corColors[4]
  } else {
    corCol <- corColors[5]
  }

  p <- p +
    theme(panel.background = element_rect(fill = corCol))

  return(p)
}

plotMAWithLabels <- function(results, main, alpha = 0.05) {
  
  # Plot MA plot with lamin genes highlighted 
  plotMA(results, ylim = c(-4, 4), main = main,
         ylab = "log2 fold change", alpha = alpha)
  points(results[c("ENSG00000165029", "ENSG00000160179"), "baseMean"],
         results[c("ENSG00000165029", "ENSG00000160179"), "log2FoldChange"],
         col = c("red"), cex = 1.5, pch = 19)
  legend("topright", 
         legend = c(paste0("non-sign (n=", sum(results$padj >= alpha, na.rm = T), ")"), 
                    paste0("sign (n=", sum(results$padj < alpha, na.rm = T), ")"), 
                    "ABCA1"), 
         col = c("black", "blue", "red"), pch = 19)
  
}

```


### Load count table

Read data files.

```{r load rna-seq}
## Load gene counts
gene_counts <- read_delim("/shared/gcf/m.trauernicht/7813/genecounts.txt") %>%
  filter(chromosome_name %in% c(paste0(1:22), "X")) %>%
  filter(gene_biotype == "protein_coding") %>%
  dplyr::select("tf" = external_gene_id, contains("mES")) %>%
  mutate(tf = toupper(tf)) %>%
  pivot_longer(-tf, names_to = "sample", values_to = "count") %>%
  # Remove numbers in front of sample names
  mutate(sample = gsub("[0-9]{1}_", "", sample)) %>%
  mutate(count = ave(count, tf, sample, FUN = mean)) %>%
  mutate(count = round(count, 0)) %>%
  distinct() %>%
  mutate(replicate = gsub(".*_(r[1-3]{1}$)", "\\1", sample)) %>%
  mutate(condition = gsub("(.*)_r[1-3]{1}$", "\\1", sample))

gene_count_mat <- gene_counts %>%
  dplyr::select(-replicate,-condition) %>%
  spread(sample, count) %>%
  column_to_rownames("tf") %>%
  replace(is.na(.), 0)
```

---

## Compute VIPER TF activity

```{r}
# acessing (mouse) dorothea regulons
dorothea_mm <- decoupleR::get_dorothea(organism='mouse', levels=c('A', 'B', 'C'))

dorothea_mm <- dorothea_mm %>%
  dplyr::rename("tf"="source") %>%
  mutate(tf = toupper(tf)) %>%
  mutate(target = toupper(target))

# run viper
tf_activities <- dorothea::run_viper(gene_count_mat, 
                                      dorothea_mm,
                                      options = list(minsize = 4,
                                                     method = "ttest"))

tf_activities_viper <- tf_activities %>%
  as.data.frame() %>%
  rownames_to_column("tf") %>%
  pivot_longer(-tf, names_to = "sample", values_to = "viper_activity")
  
saveRDS(tf_activities_viper, "/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/RNA_seq_analyses/tf_activities_viper.rds")


## Check why some TFs go up (e.g. HNF1A)
dorothea_mm_ex <- dorothea_mm %>%
  filter(tf == "SOX2")

gene_count_mat_ex <- gene_count_mat %>%
  rownames_to_column("tf") %>%
  filter(tf %in% dorothea_mm_ex$target) %>%
  pivot_longer(-tf, names_to = "sample", values_to = "count") %>%
  mutate(condition = gsub("_r[1-3]{1}$", "", sample)) %>%
  mutate(replicate = gsub(".*_(r[1-3]{1}$)", "\\1", sample)) %>%
  distinct(condition, tf, replicate, count) %>%
  #mutate(count = log2(count)) %>%
  spread(condition, count)

ggplot(gene_count_mat_ex,
       aes(x = `mES_2i`, y = `mES_2i_LIF`)) +
  geom_abline() +
  geom_point() +
  theme_pubr() +
  labs_pubr()


```

## Import ISMARA activities
```{r}
metadata_df <- gene_counts %>%
  distinct(sample, condition, replicate)

ismara_activities <- read_delim("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/RNA_seq_analyses/activity_table") %>%
  dplyr::rename("sample" = "#sample") %>%
  pivot_longer(-sample, names_to = "tf", values_to = "ismara_activity") %>%
  mutate(tf = toupper(tf)) %>%
  mutate(sample = gsub("7813_[1-6]_(.*r[1-3])_.*", "\\1", sample)) %>%
  left_join(metadata_df, by = "sample") %>%
  mutate(mean_ismara_activity = ave(ismara_activity, tf, condition, FUN = mean))
  
ref_ismara <- ismara_activities %>%
  filter(condition == "mES_2i_LIF") %>%
  dplyr::select(tf, "ref_ismara_activity" = ismara_activity) %>%
  distinct()

ismara_activities <- ismara_activities %>%
  left_join(ref_ismara, by = "tf") %>%
  mutate(ismara_activity_dif = ismara_activity / ref_ismara_activity) %>%
  mutate(mean_ismara_activity_dif = ave(ismara_activity_dif, tf, FUN = mean))


saveRDS(ismara_activities, "/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/RNA_seq_analyses/ismara_activities.rds")
```


## 1. PCA plot

```{r initialize deseq2, fig.width = 6, fig.height = 3.5}

#######################################
## Prepare count matrix

# Convert into data.frame with row.names for deseq2
rnaseq_counts <- gene_counts %>%
  distinct(tf, count, sample) %>%
  spread(key = "sample", value = "count") %>%
  replace(is.na(.), 0) %>%
  column_to_rownames("tf")


#######################################
## Initialize deseq2

# Prepare metadata ready for deseq
metadata_df <- gene_counts %>%
  distinct(sample, condition, replicate) %>%
  mutate(ctrl = ifelse(condition == "mES_2i_LIF", "yes", "no"))

# Initialize
rnaseq_dds <- DESeqDataSetFromMatrix(countData = rnaseq_counts,
                                     colData = metadata_df,
                                     design= ~ condition)

# Execute deseq2
rnaseq_dds <- DESeq(rnaseq_dds)


#######################################
## PCA analysis - quality control

# Get the "normalized" values and create PCA plot
# Note that "normalized" is simply log2 + 0.01 transformed normalized counts!
rnaseq_dds_norm <- normTransform(rnaseq_dds, pc = 0.01)
plt <- plotPCA(rnaseq_dds_norm, intgroup = c("condition"), ntop = 100)

plt + 
  theme_pubr() +
  labs_pubr() +
  theme(aspect.ratio = 1) +
  scale_shape_manual()
```

---

## 2. Correlation plot replicates

A quick initial test: how well do the replicates correlate with each other and
between the different experiments?

```{r correlation plots, fig.width = 6, fig.height = 6, message = F, warning = F}

#######################################
## Make a correlation plot of all samples - quality control

# Get the "normalized" counts for each experiment - this is a different
# normalized from above!
rnaseq_norm <- data.frame(assay(rnaseq_dds_norm)) %>%
  rownames_to_column("tf")

# Get a sample for plotting dplyr::select points
n <- sample(1:nrow(rnaseq_norm), 1000)

# Use GGally to make correlation plots
boundaries <- seq(from = 0.9, by = 0.01, length.out = 4)
plt <- ggpairs(rnaseq_norm %>% dplyr::select(-tf),
               upper = list(continuous = corColor),
               lower = list(continuous = function(data, mapping, ...) {
                   ggally_points(data = data[n, ], mapping = mapping, alpha = 0.1, size = 0.5) +
                   geom_abline(slope = 1, lty = "dashed", col = "red") +
                   theme_bw()}),
               diag = list(continuous = function(data, mapping, ...) {
                   ggally_densityDiag(data = data, mapping = mapping, alpha = 0.3, fill = "red") +
                   theme_bw()})) +
  ggtitle("Correlation gene expression") +
  xlab("Gene expression") +
  ylab("Gene expression")
  # theme_bw()

print(plt)

```



---


## 3. Clustering of correlation matrix

Are the replicates clustering together?
```{r correlation matrix, fig.width = 5.5, fig.height = 5}
norm <- assay(vst(rnaseq_dds, nsub = 10))
dist <- as.dist(1 - cor(norm, method = "spearman"))

clust <- hclust(dist, method = "ward.D2")
clust <- reorder(clust, dist = dist, method = "OLO")

df <- reshape2::melt(cor(norm, method = 'spearman'))

ggplot(df, aes(Var1, Var2, fill = value)) +
  geom_raster() +
  guides(y.sec = "axis", x.sec = guide_axis(position = "bottom")) +
  GENOVA:::scale_fill_GENOVA_div(name = expression("Spearman's "*rho)) +
  theme(axis.text.x.bottom = element_text(angle = 90, hjust = 1, vjust = 0.5))
```


---

## 4. Differential analysis

Identifying individual genes that change expression upon stimulation. 
  
```{r differential analysis, fig.width = 5.5, fig.height = 5}

#######################################
## Differential analysis - test for lfc > 0.5

# Differential analysis
diff_results <- results(rnaseq_dds,
                       contrast = c("condition", "mES_2i", "mES_2i_LIF"))

plotCounts(rnaseq_dds, gene="STAT3", intgroup="condition")

# MA plots with significant hits
ggmaplot(diff_results, main = expression("mES_2i_LIF" %->% "mES_2i"),
   fdr = 0.05, fc = 1.5, size = 1,
   palette = c("#B31B21", "#1465AC", "darkgray"),
   genenames = as.vector(diff_results$name),
   legend = "top", top = 100,
   font.label = c("bold", 11),
   font.legend = "bold",
   font.main = "bold",
   ggtheme = ggpubr::theme_pubr())


# MA plot of only STAT3 target genes
dorothea_stat3 <- decoupleR::get_dorothea(organism='mouse', levels=c('A', 'B', 'C'))

dorothea_target <- dorothea_mm %>%
  filter(tf == "KLF4")

targets <- unique(dorothea_target$target)

# Subset the DESeqDataSet by row names
diff_results_stat3_targets <- diff_results[rownames(diff_results) %in% targets, ]

ggmaplot(diff_results_stat3_targets, main = expression("mES_2i_LIF" %->% "mES_2i"),
   fdr = 0.05, fc = 1.5, size = 1,
   palette = c("#B31B21", "#1465AC", "darkgray"),
   genenames = as.vector(diff_results$name),
   legend = "top", top = 100,
   font.label = c("bold", 11),
   font.legend = "bold",
   font.main = "bold",
   ggtheme = ggpubr::theme_pubr())

# Extract hits from all stimulations
df_changed_LIF <- data.frame("tf" = rownames(diff_results),
                         "basal_exp" = diff_results$baseMean,
                         "lfc" = diff_results$log2FoldChange, 
                         "p_value" = diff_results$pvalue) %>%
  mutate(stat3_target = ifelse(tf %in% targets, "yes", "no")) %>%
  mutate(stat3_target = ifelse(tf == "KLF4", "klf4", stat3_target))

ggplot(,mapping = aes(x = lfc, y = -log10(p_value))) +
  geom_point(data = df_changed_LIF %>% filter(stat3_target == "no", (p_value >= .05 | abs(lfc) < 1)), color = "grey", alpha = .2, stroke = NA, size = 2) +
  geom_point(data = df_changed_LIF %>% filter(stat3_target == "no", (p_value < .05 & abs(lfc) >= 1)), color = "grey", alpha = .7, stroke = NA, size = 2) +
  geom_point(data = df_changed_LIF %>% filter(stat3_target == "yes", (p_value >= .05 | abs(lfc) < 1)), color = "red", alpha = .2, stroke = NA, size = 2) +
  geom_point(data = df_changed_LIF %>% filter(stat3_target == "yes", (p_value < .05 & abs(lfc) >= 1)), color = "red", stroke = NA, size = 2) +
  geom_point(data = df_changed_LIF %>% filter(stat3_target == "klf4", (p_value >= .05 | abs(lfc) < 1)), color = "blue", alpha = .2, stroke = NA, size = 1.5) +
  geom_point(data = df_changed_LIF %>% filter(stat3_target == "klf4", (p_value < .05 & abs(lfc) >= 1)), color = "blue", stroke = NA, size = 2) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = -1, linetype = "dashed") +
  theme_pubr() +
  theme(legend.position = "top") +
  labs(x = "log2 fold change", y = "-log10 p-value", color = "STAT3 target gene") +
  labs_pubr()

saveRDS(df_changed_LIF, "/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/RNA_seq_analyses/deseq_changed_LIF.rds")
```

---

### Conclusion



---

### SessionInfo

```{r sessioninfo}

sessionInfo()

```
