---
title: "Comparison of ATAC-seq inferred activity with TF reporter activity"
author: 
  - name: "Max Trauernicht"
    email: "m.trauernicht@nki.nl"
    affiliation: "Netherlands Cancer Institute"
date: '20/02/2025'
output: 
  html_document:
    theme: flatly  # Modern, professional look
    highlight: tango  # Clean code highlighting
    fig_caption: true
    code_folding: show  # Users can expand/collapse code
    df_print: paged  # Paginate long tables instead of cluttering the page
    toc: true
    toc_depth: 2  # Limits TOC depth for clarity
    toc_float: 
      collapsed: true  # Keeps TOC clean, expandable when needed
      smooth_scroll: true  # Smooth scrolling when clicking TOC links
    number_sections: true  # Automatically number sections
    mathjax: "default"  # Enables LaTeX-style equations if needed
---

---
 
# Aim
Comparing activities between ATAC-seq inferred TF activities and activities measured using optimized prime TF reporters.

---
 
# Setup

---

## Libraries 

```{r setup, out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
knitr::opts_chunk$set(echo = TRUE)
StartTime <-Sys.time()

# 8-digit Date tag:
Date <- substr(gsub("-","",Sys.time()),1,8) 
# libraries:
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(tibble)
library(pheatmap)
library(visNetwork)
library(ggbeeswarm)
library(ggforce)
library(viridis)
library(purrr)
library(plyr)
library(igraph)
library(ggraph)
library(cowplot)
library(gridExtra)
library(pROC)
library(tidyr)
library(stringr)
library(ggrastr)
library(readr)
library(ggbiplot)
library(ggh4x)
library(ggiraph)
library(plotly)
library(AMR)
library(umap)
library(ggrepel)
library(chromVAR)
library(Matrix)
library(SummarizedExperiment)
library(BiocParallel)
library(BSgenome.Mmusculus.UCSC.mm10)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ggnewscale)
```

---

## Functions

```{r out.width= "80%", fig.align= "center"}
### Custom functions
SetFileName <- function(filename, initials) {
  # Set filename with extension and initials to make filename with date integrated.
  filename <- substitute(filename)
  initials <- substitute(initials)
  filename <- paste0(initials, Date, filename)
  filename
}


# Extract p-value from linear model
lmp <- function (modelobject) {
    if (class(modelobject) != "lm") stop("Not an object of class 'lm' ")
    f <- summary(modelobject)$fstatistic
    p <- pf(f[1],f[2],f[3],lower.tail=F)
    attributes(p) <- NULL
    return(p)
}

theme_custom <- theme(
  text = element_text(size = 12),
  
  # Remove panel background
  panel.background = element_blank(),
  
  # Customize grid lines: Keep both major X and Y grid lines, make them thinner
  panel.grid = element_blank(),
  panel.grid.major = element_line(colour = "#e3e1e1", linetype = 1, size = 0.3),
  
  # Ensure axis lines and ticks are visible
  axis.line = element_line(colour = "black"),
  axis.ticks = element_line(colour = "black"),
  
  # Make axis ticks longer
  axis.ticks.length = unit(7, "pt"),  # Adjust the length as needed
  
  # Add an outside border to the plotting area
  panel.border = element_rect(colour = "black", fill = NA, size = .5),
  
  # Ensure facet (strip) outline matches the panel border thickness
  strip.background = element_rect(fill = "grey90", colour = "black", size = .5),
  strip.text = element_text(face = "bold"),
  
  # Adjust plot layout
  plot.title.position = "plot",
  legend.position = "top",
  legend.title = element_blank()
)

theme_set(theme_minimal() + theme_custom)

# Function to force axes to start at zero
force_zero_axes <- function(p) {
  p + 
    scale_x_continuous(expand = c(0, 0), limits = c(0, NA)) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, NA))
}
```

---

## Load pre-computed TF reporter and chromVAR data
```{r out.width= "80%", fig.align= "center"}
prime_activities_comb <- read_csv("/DATA/shared/projects/TF_REPORTERS/ATAC_TF_reporter_comparison/TF_reporter_analyses/data/mt20250220_prime_reporter_activities_combined.csv")

chromvar_scores_mouse_added_all <- readRDS("/DATA/shared/projects/TF_REPORTERS/ATAC_TF_reporter_comparison/data/chromVAR/dev_scores_mouse_2025b_mouse_all_npc.rds")
```

---

# Data visualization

---

## Plot chromVAR z-scores upon SOX2 degradation

```{r out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
chromvar_scores_mouse_added_sox2 <- chromvar_scores_mouse_added_all %>%
  filter(condition %in% c("mES_SOX2_DEG")) %>%
  distinct(tf, condition, chromVAR_zscore_mean, p_value_adjusted, ctrl_zscore) %>%
  mutate(ctrl_zscore = ave(ctrl_zscore, tf, FUN = function(x) mean(x))) %>%
  distinct() %>%
  mutate(
    sig = case_when(
      p_value_adjusted < 0.05 & ctrl_zscore > chromVAR_zscore_mean ~ "Downregulated",
      p_value_adjusted < 0.05 & ctrl_zscore < chromVAR_zscore_mean ~ "Upregulated", TRUE ~ "NS"), 
    sig = factor(sig, levels = c("NS", "Downregulated", "Upregulated"))) %>%
  mutate(color_axis = ifelse(sig == "Upregulated", "#f37f80", ifelse(sig == "NS", "gray30", "#6495ed"))) %>%
  arrange(desc(ctrl_zscore)) %>%
  mutate(tf = factor(tf, levels = unique(tf))) %>%
  distinct(tf, .keep_all = T) %>%
  arrange(desc(ctrl_zscore)) %>%
  mutate(tf = factor(tf, levels = unique(tf)))


chromvar_scores_mouse_added_sox2 %>%
            ggplot() +
            geom_segment(aes(
                y = ctrl_zscore, 
                yend = chromVAR_zscore_mean,
                x = tf, xend = tf,
                color = sig
            ), size = 1) +
            scale_color_manual(
                values = c("NS" = "grey", "Downregulated" = "#6495ed", "Upregulated" = "#f37f80"),
                name = "Result of\nComparative\nAnalysis"
            ) +
            new_scale_color() +
            geom_point(aes(x = tf, y = ctrl_zscore, color = "C"), size = 3) +
            geom_point(aes(x = tf, y = chromVAR_zscore_mean, color = sig), size = 3) +
            scale_color_manual(values = c('C'= "black", "NS" = "grey", "Downregulated" = "#6495ed", "Upregulated" = "#f37f80")) +
            guides(color = "none") +
            theme_bw() +
            theme(
                axis.text.x = element_text(
                    angle = 90, hjust = 1, vjust = 0.5,
                    color = chromvar_scores_mouse_added_sox2$color_axis
                ),
                # Remove inner lines of y-axis
                panel.grid.major.y = element_blank(),
                panel.grid.minor.y = element_blank(),
                text = element_text(size = 14)
            ) +
            labs(
                y = "Activity (log2(RPM+1))",
                x = "",
                title= paste("mES_SOX2_DEG", "vs.", "mES_SOX2_ctrl")
            )


chromvar_scores_mouse_added_sox2 <- chromvar_scores_mouse_added_all %>%
  filter(condition %in% c("mES_SOX2_DEG")) %>%
  distinct(tf, condition, chromVAR_zscore_dif_mean, p_value_adjusted) %>%
  mutate(
    sig = case_when(
      p_value_adjusted < 0.05 & chromVAR_zscore_dif_mean < 0 ~ "Downregulated",
      p_value_adjusted < 0.05 & chromVAR_zscore_dif_mean > 0 ~ "Upregulated", TRUE ~ "NS"), 
    sig = factor(sig, levels = c("NS", "Downregulated", "Upregulated"))) 

ggplot(chromvar_scores_mouse_added_sox2, aes(x = chromVAR_zscore_dif_mean, y = -log10(p_value_adjusted), color = sig, alpha = sig, tf=tf)) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") +
  geom_point() +
  # sig will be tab-blue
  scale_color_manual(values = c("NS" = "grey", "Downregulated" = "#6495ed", "Upregulated" = "#f37f80")) +
  scale_alpha_manual(values = c("NS" = 0.4, "Upregulated" = 1, "Downregulated" = 1)) +
  # annotate the significant points
  geom_label_repel(data = . %>% filter(sig != "NS"), size=5, aes(label = tf), box.padding = 0.5) +
  guides(color = "none", alpha = "none") +
  theme(aspect.ratio = 1)

chromvar_scores_mouse_added_sox2 <- chromvar_scores_mouse_added_all %>%
  filter(condition %in% c("mES_SOX2_DEG"), tf == "SOX2") %>%
  distinct(chromVAR_zscore, ctrl_zscore) %>%
  pivot_longer(ctrl_zscore:chromVAR_zscore, names_to = "type", values_to = "zscore")

chromvar_scores_mouse_added_sox2_mean <- chromvar_scores_mouse_added_sox2 %>%
  group_by(type) %>%
  summarise(zscore = mean(zscore))

ggplot() +
  geom_bar(data = chromvar_scores_mouse_added_sox2_mean, aes(x = type, y = zscore), stat = "identity", fill = "grey") +
  geom_point(data = chromvar_scores_mouse_added_sox2, aes(x = type, y = zscore),  width = .1, size = 3, shape = 21, fill = "grey") +
  theme(aspect.ratio = 2.5)
```
Conclusion: Oct4::Sox2 and Sox2 go down, which is a good quality check!
---

## Correlate conditions
```{r out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
ggplot(chromvar_scores_mouse_added_all %>%
         mutate(ctrl_zscore = ave(ctrl_zscore, tf, condition, FUN = mean)) %>%
         distinct(tf, condition, chromVAR_zscore_dif_mean, ctrl_zscore) %>%
         filter(condition == "mES_POU5F1_DEG"), aes(x = chromVAR_zscore_dif_mean, y = ctrl_zscore)) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  theme(aspect.ratio = 1)


chromvar_scores_mouse_1 <- chromvar_scores_mouse_added_all %>%
  filter(condition == "mES_POU5F1_DEG") %>%
  dplyr::select(tf, "chromVAR_1" = chromVAR_zscore_mean) %>%
  distinct()

chromvar_scores_mouse_2 <- chromvar_scores_mouse_added_all %>%
  filter(condition == "mES_SOX2_DEG") %>%
  dplyr::select(tf, "chromVAR_2" = chromVAR_zscore_mean) %>%
  distinct()

chromvar_scores_mouse_comb <- chromvar_scores_mouse_1 %>%
  left_join(chromvar_scores_mouse_2, by = "tf")

ggplot(chromvar_scores_mouse_comb, aes(x = chromVAR_1, y = chromVAR_2)) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  theme(aspect.ratio = 1)
```
Conclusion: ChromVar replicates do correlate.

---

## Correlate inferred activity with measured activity - steady state
```{r out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
## Plot inferred TF activity in mESCs
ggplot(chromvar_scores_mouse_added_all %>% 
         filter(condition == "mES_FK") %>%
         mutate(ctrl_zscore = ave(ctrl_zscore, tf, condition, FUN = function(x) mean(x, na.rm = T))),
       aes(x = reorder(tf, -ctrl_zscore), y = ctrl_zscore)) +
  geom_bar(stat = "identity") +
  xlab("TF") +
  ylab("chromVAR z-score") +
  theme(aspect.ratio = 1/4, 
        axis.text.x = element_text(angle = 90, hjust = 1)) 

ggplot(chromvar_scores_mouse_added_all %>% 
         filter(condition == "mES_FK") %>%
         mutate(ctrl_zscore = ave(ctrl_zscore, tf, condition, FUN = function(x) mean(x, na.rm = T))) %>%
         left_join(prime_activities_comb %>% 
                     filter(condition == "mES_FK") %>% 
                     mutate(ctrl_activity = ave(ctrl_activity, tf, condition, FUN = function(x) mean(x, na.rm = T))) %>%
                     distinct(tf, ctrl_activity)) %>%
         distinct(tf, ctrl_zscore, ctrl_activity),
       aes(x = ctrl_zscore, y = ctrl_activity)) +
  geom_abline(lty = 2) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point() +
  xlab("chromVAR z-score") +
  ylab("TF reporter activity (log2)") +
  theme(aspect.ratio = 1) 
```
Conclusion: Not much correlation on steady-state level. We should probably look at the fold-changes.

---

## Replicate correlation FC - reporter activity
```{r out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
## Correlate chromVAR activities with measured activity
tf_activity_chromVAR_dev <- prime_activities_comb %>%
  distinct(tf, logFC, replicate, condition)

## Correlation between replicates within assays
tf_activity_chromVAR_dev_2 <- tf_activity_chromVAR_dev %>%
  spread(replicate, logFC)

ggplot(tf_activity_chromVAR_dev_2, aes(x = R1, y = R2)) +
  geom_smooth(method = "lm", color = "red") +
  geom_hline(yintercept = 0, lty = 1) +
  geom_vline(xintercept = 0, lty = 1) +
  geom_abline(lty = 2) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point(size = 3, shape = 21, fill = "grey") +
  labs(x = "Reporter activity (FC) R1", y = "Reporter activity (FC) R2") +
  facet_wrap(~condition)

ggplot(tf_activity_chromVAR_dev_2 %>% filter(condition == "mES_SOX2_DEG"), aes(x = R1, y = R2)) +
  geom_smooth(method = "lm", color = "red") +
  geom_hline(yintercept = 0, lty = 1) +
  geom_vline(xintercept = 0, lty = 1) +
  geom_abline(lty = 2) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point(size = 3, shape = 21, fill = "grey") +
  labs(x = "Reporter activity (FC) R1", y = "Reporter activity (FC) R2") +
  facet_wrap(~condition) +
  theme(aspect.ratio = 1)

```
Conclusion: Reporter activity fold-changes do correlate quite well!

---

## Replicate correlation FC - chromVAR
```{r out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
tf_activity_chromVAR_dev_3 <- chromvar_scores_mouse_added_all %>%
  distinct(condition, tf, chromVAR_zscore_dif, replicate) %>%
  spread(replicate, chromVAR_zscore_dif)

ggplot(tf_activity_chromVAR_dev_3, aes(x = R1, y = R2)) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point() +
  geom_smooth(method = "lm", color = "red") +
  geom_abline(lty = 2) +
  labs(x = "Motif accessibility (FC) R1", y = "Motif accessibility (FC) R2") +
  facet_wrap(~condition)

ggplot(tf_activity_chromVAR_dev_3 %>% filter(condition == "mES_SOX2_DEG"), aes(x = R1, y = R2)) +
  geom_smooth(method = "lm", color = "red") +
  geom_hline(yintercept = 0, lty = 1) +
  geom_vline(xintercept = 0, lty = 1) +
  geom_abline(lty = 2) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point(size = 3, shape = 21, fill = "grey") +
  labs(x = "Motif accessibility (FC) R1", y = "Motif accessibility (FC) R2") +
  theme(aspect.ratio = 1)

## Correlation heatmap
tf_activity_chromVAR_dev_3 <- chromvar_scores_mouse_added_all %>%
  filter(condition %in% c("mES_SOX2_DEG")) %>%
  distinct(tf, chromVAR_zscore_mean, replicate, ctrl_zscore) %>%
  pivot_longer(cols = c(chromVAR_zscore_mean, ctrl_zscore),
               names_to = "condition",
               values_to = "chromVAR_score") %>%
  mutate(sample = paste(condition, replicate, sep = "_")) %>%
  dplyr::select(-replicate, -condition) %>%
  spread(sample, chromVAR_score) %>%
  column_to_rownames("tf") %>%
  as.matrix()

tf_activity_chromVAR_dev_cor <- cor(tf_activity_chromVAR_dev_3, method = "pearson", use = "pairwise.complete.obs")

## Correlation heatmap
pheatmap(tf_activity_chromVAR_dev_cor,
         border = "white",
         cellwidth = 10,
         cellheight = 10)

```
Conclusion: chromVAR fold-changes correlate pretty well between replicates.

---

## Reporter activity of SOX2 reporters upon SOX2 degradation
```{r out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
## Barcode counts SOX2 reporters upon SOX2 degradation
## Load data
primetime_sox2_bc <- read_delim("~/projects/prime_TF_reporters/mES_ATAC_primetime/01_OUTPUT/mES_SOX2/tmp_primetime/activity/cDNA_counts.txt") %>%
  pivot_longer(cols = c(-tf, -negative_control, -promoter, -barcode),
               names_to = "sample",
               values_to = "reads") %>%
  mutate(rpm = ave(reads, sample, FUN = function(x) (x) / sum(x) *1e6 )) %>%
  filter(tf == "SOX2") %>%
  mutate(condition = gsub("_[1-3]{1}", "", sample))

primetime_pDNA <- read_delim("~/projects/prime_TF_reporters/mES_ATAC_primetime/01_OUTPUT/mES_SOX2/tmp_primetime/activity/pDNA_counts.txt") %>%
    mutate(pDNA_rpm = pDNA / sum(pDNA) *1e6 )
  
primetime_sox2_bc <- primetime_sox2_bc %>%
  left_join(primetime_pDNA) %>%
  mutate(activity = rpm / pDNA_rpm) 

primetime_sox2_bc_mean <- primetime_sox2_bc %>%
  mutate(rpm_mean = ave(rpm, condition, FUN = function(x) mean(x, na.rm = T))) %>%
  distinct(rpm_mean, condition)

ggplot() +
  geom_bar(data = primetime_sox2_bc_mean, aes(x = condition, y = rpm_mean), stat = "identity", fill = "grey") +
  geom_jitter(data = primetime_sox2_bc, aes(x = condition, y = rpm), width = .1, size = 3, shape = 21, fill = "grey") +
  theme(aspect.ratio = 2.5)

## Correlate chromVAR activities with measured activity
tf_activity_chromVAR_dev <- prime_activities_comb %>%
  mutate(condition = ifelse(condition == "mES_LIF1", "mES_LIF_1h", condition)) %>%
  mutate(condition = ifelse(condition == "mES_LIF6", "mES_LIF_6h", condition)) %>%
  mutate(condition = ifelse(condition == "mES_LIF3", "mES_LIF_3h", condition)) %>%
  left_join(chromvar_scores_mouse_added_all %>% 
              mutate(chromVAR_zscore_dif = ave(chromVAR_zscore_dif, tf, condition, FUN = function(x) mean(x, na.rm = T))) %>%
              distinct(tf, condition, chromVAR_zscore_dif) %>%
              mutate(condition = ifelse(condition == "NPC_DMSO", "NPC", condition)) %>%
              mutate(condition = ifelse(condition == "mES_2i_LIF", "mES_NPC", condition))) %>%
  distinct(tf, logFC, chromVAR_zscore_dif, condition, sig, ctrl_activity) %>%
  mutate(logFC = ave(logFC, tf, condition, FUN = function(x) mean(x, na.rm = T))) %>%
  distinct()

ggplot(tf_activity_chromVAR_dev %>%
         mutate(pert_activity = ctrl_activity + logFC) %>%
         filter(tf == "SOX2", condition %in% c("mES_SOX2_DEG")) %>%
         distinct(pert_activity, ctrl_activity) %>%
         pivot_longer(cols = c(pert_activity, ctrl_activity),
                      values_to = "activity",
                      names_to = "condition"),
       aes(x = condition, y = activity)) +
  geom_bar(stat = "identity", fill = "grey") +
  geom_point(size = 3, shape = 21, fill = "grey") +
  theme(aspect.ratio = 2.5)
```
Conclusion: Sox2 reporter activity drops upon Sox2 depletion.

---

## Correlate primetime FCs with chromVAR FCs
```{r out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
tf_activity_chromVAR_dev <- prime_activities_comb %>%
  mutate(condition = ifelse(condition == "mES_LIF1", "mES_LIF_1h", condition)) %>%
  mutate(condition = ifelse(condition == "mES_LIF6", "mES_LIF_6h", condition)) %>%
  mutate(condition = ifelse(condition == "mES_LIF3", "mES_LIF_3h", condition)) %>%
  mutate(condition = ifelse(condition == "mES_NPC", "NPC_DMSO", condition)) %>%
  filter(!condition %in% c("mES_PMA", "mES_SP1", "NPC_FK", "NPC_HQ")) %>%
  left_join(chromvar_scores_mouse_added_all %>% 
              mutate(chromVAR_zscore_dif_mean = ifelse(condition == "mES_FOXA1_OE", chromVAR_zscore_dif, chromVAR_zscore_dif_mean)) %>%
              distinct(tf, condition, chromVAR_zscore_dif_mean, p_value_adjusted)) %>%
  distinct(tf, logFC, chromVAR_zscore_dif_mean, condition, sig, p_adjusted, p_value_adjusted) %>%
  mutate(logFC = ave(logFC, tf, condition, FUN = function(x) mean(x, na.rm = T))) %>%
  distinct() %>%
  mutate(sig_chrom = ifelse(p_value_adjusted < 0.05 & chromVAR_zscore_dif_mean > 0, "Upregulated", "NS"),
         sig_chrom = ifelse(p_value_adjusted < 0.05 & chromVAR_zscore_dif_mean < 0, "Downregulated", sig_chrom))


ggplot(tf_activity_chromVAR_dev %>%
         filter(!condition %in% c("mES_LIF_1h", "mES_LIF_3h", "mES_LIF_6h", "NPC_DMSO")) %>%
         mutate(condition = factor(condition, levels = c("mES_CH", "mES_LIF", "mES_SOX2_DEG",
                                                         "mES_POU5F1_DEG", "mES_TFCP2L1", "mES_FBS",
                                                         "mES_FK", "mES_FOXA1_OE", "mES_Heat")),
                x_sig = !is.na(p_value_adjusted) & p_value_adjusted < 0.05,
                y_sig = sig != "NS",
                low_sig = !(x_sig | y_sig)),  # insignificant for both metrics
       aes(x = chromVAR_zscore_dif_mean, y = logFC)) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  # Dots with NO outline (x_sig = FALSE)
  geom_point(data = ~ subset(., !x_sig),
             aes(fill = sig, alpha = low_sig),
             shape = 21, size = 3, stroke = 0) +
  # Dots with black outline (x_sig = TRUE)
  geom_point(data = ~ subset(., x_sig),
             aes(fill = sig, alpha = low_sig),
             shape = 21, size = 3, stroke = .7, color = "black") +
  geom_smooth(method = "lm", color = "black") +
  labs(x = "chromVAR deviation z-score dif", y = "Reporter activity (FC)") +
  scale_fill_manual(values = c("Downregulated" = "#6495ED", 
                               "NS" = "grey80", 
                               "Upregulated" = "#F37F80")) +
  scale_alpha_manual(values = c(`TRUE` = 0.5, `FALSE` = 1), guide = "none") +
  theme(aspect.ratio = 1) +
  facet_wrap(~condition)

## mES - NPC comparison
prime_reporters <- read_csv("/DATA/shared/projects/TF_REPORTERS/SuRE-TF/data/best_reporters.csv") %>%
  mutate(reporter_id = ifelse(tf == "POU5F1::SOX2", "POU5F1::SOX2_5bp_21bp_minP_1", reporter_id)) 
prime_tfs <- prime_reporters$tf[prime_reporters$conf_level > 1]

ggplot(tf_activity_chromVAR_dev %>%
         filter(condition %in% c("NPC_DMSO")) %>%
         mutate(x_sig = !is.na(p_value_adjusted) & p_value_adjusted < 0.05,
                y_sig = sig != "NS",
                low_sig = !(x_sig | y_sig)),  # insignificant for both metrics
       aes(x = chromVAR_zscore_dif_mean, y = logFC)) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  # Dots with NO outline (x_sig = FALSE)
  geom_point(data = ~ subset(., !x_sig),
             aes(fill = sig, alpha = low_sig),
             shape = 21, size = 3, stroke = 0) +
  # Dots with black outline (x_sig = TRUE)
  geom_point(data = ~ subset(., x_sig),
             aes(fill = sig, alpha = low_sig),
             shape = 21, size = 3, stroke = .7, color = "black") +
  geom_smooth(method = "lm", color = "black") +
  labs(x = "chromVAR deviation z-score dif", y = "Reporter activity (FC)") +
  scale_fill_manual(values = c("Downregulated" = "#6495ED", 
                               "NS" = "grey80", 
                               "Upregulated" = "#F37F80")) +
  scale_alpha_manual(values = c(`TRUE` = 0.5, `FALSE` = 1), guide = "none") +
  theme(aspect.ratio = 1) +
  xlim(-0.5, 1) +
  ylim(-5,6)

cor(tf_activity_chromVAR_dev$logFC[tf_activity_chromVAR_dev$condition == "NPC_DMSO" & tf_activity_chromVAR_dev$tf %in% prime_tfs], 
    tf_activity_chromVAR_dev$chromVAR_zscore_dif_mean[tf_activity_chromVAR_dev$condition == "NPC_DMSO" & tf_activity_chromVAR_dev$tf %in% prime_tfs], 
    use = "pairwise.complete.obs", method = "spearman")

### Only pluripotency TFs
ggplot(tf_activity_chromVAR_dev %>% 
         filter(tf %in% c("POU5F1::SOX2", "SOX2", "TCF7",
                          "STAT3", "KLF4", "TFCP2L1", "POU5F1",
                          "ESRRB", "SMAD4"),
                condition == "NPC_DMSO") %>%
         pivot_longer(cols = c(logFC, chromVAR_zscore_dif_mean),
               values_to = "activity",
               names_to = "method") %>%
         filter(method == "logFC"), aes(x = reorder(tf, -activity), y = activity, fill = sig)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +  # Add a vertical line at x = 0
  coord_flip() +  # Flip the plot to make it horizontal
  scale_y_continuous(labels = abs) +  # Display positive labels for both sides
  labs(x = "Category", y = "Value", title = "Double-Sided Bar Plot") +
  scale_fill_manual(values = c("Upregulated" = "#F37F80", "NS" = "grey80", "Downregulated" = "#6495ED")) +
  theme(aspect.ratio = 2) +
  ylim(-6, 6)

ggplot(tf_activity_chromVAR_dev %>% 
         filter(tf %in% c("POU5F1::SOX2", "SOX2", "TCF7",
                          "STAT3", "KLF4", "TFCP2L1", "POU5F1",
                          "ESRRB", "SMAD4"),
                condition == "NPC_DMSO") %>%
         pivot_longer(cols = c(logFC, chromVAR_zscore_dif_mean),
               values_to = "activity",
               names_to = "method") %>%
         filter(method == "chromVAR_zscore_dif_mean") %>%
         mutate(tf = factor(tf, levels = rev(c("TCF7", "STAT3", "SOX2", "TFCP2L1",
                                           "KLF4", "POU5F1::SOX2", "SMAD4", "ESRRB",
                                           "POU5F1")))), aes(x = tf, y = activity, fill = sig_chrom)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +  # Add a vertical line at x = 0
  coord_flip() +  # Flip the plot to make it horizontal
  scale_y_continuous(labels = abs) +  # Display positive labels for both sides
  labs(x = "Category", y = "Value", title = "Double-Sided Bar Plot") +
  scale_fill_manual(values = c("Upregulated" = "#F37F80", "NS" = "grey80", "Downregulated" = "#6495ED")) +
  theme(aspect.ratio = 2) +
  ylim(-0.3, 0.3)



## STAT3 time-course
ggplot(tf_activity_chromVAR_dev %>%
         filter(condition %in% c("mES_LIF_1h", "mES_LIF_3h", "mES_LIF_6h", "mES_LIF")) %>%
         mutate(condition = factor(condition, levels = c("mES_LIF_1h", "mES_LIF_3h", "mES_LIF_6h", "mES_LIF")),
                x_sig = !is.na(p_value_adjusted) & p_value_adjusted < 0.05,
                y_sig = sig != "NS",
                low_sig = !(x_sig | y_sig)),  # insignificant for both metrics
       aes(x = chromVAR_zscore_dif_mean, y = logFC)) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  # Dots with NO outline (x_sig = FALSE)
  geom_point(data = ~ subset(., !x_sig),
             aes(fill = sig, alpha = low_sig),
             shape = 21, size = 3, stroke = 0) +
  # Dots with black outline (x_sig = TRUE)
  geom_point(data = ~ subset(., x_sig),
             aes(fill = sig, alpha = low_sig),
             shape = 21, size = 3, stroke = .7, color = "black") +
  geom_smooth(method = "lm", color = "black") +
  labs(x = "chromVAR deviation z-score dif", y = "Reporter activity (FC)") +
  scale_fill_manual(values = c("Downregulated" = "#6495ED", 
                               "NS" = "grey80", 
                               "Upregulated" = "#F37F80")) +
  scale_alpha_manual(values = c(`TRUE` = 0.5, `FALSE` = 1), guide = "none") +
  theme(aspect.ratio = 1) +
  facet_wrap(~condition)



for (i in unique(tf_activity_chromVAR_dev$condition)) {
  
  print(i)
  
  p <- ggplot(tf_activity_chromVAR_dev %>%
                mutate(condition = as.factor(condition),
                       x_sig = !is.na(p_value_adjusted) & p_value_adjusted < 0.05,
                       y_sig = sig != "NS",
                       low_sig = !(x_sig | y_sig)) %>%
                filter(condition == i),  # insignificant for both metrics
              aes(x = chromVAR_zscore_dif_mean, y = logFC)) +
    geom_text_repel(aes(label = tf), box.padding = 0.5) +
    # Dots with NO outline (x_sig = FALSE)
    geom_point(data = ~ subset(., !x_sig),
               aes(fill = sig, alpha = low_sig),
               shape = 21, size = 3, stroke = 0) +
    # Dots with black outline (x_sig = TRUE)
    geom_point(data = ~ subset(., x_sig),
               aes(fill = sig, alpha = low_sig),
               shape = 21, size = 3, stroke = 1.2, color = "black") +
    geom_smooth(method = "lm", color = "black") +
    labs(x = "chromVAR deviation z-score dif", y = "Reporter activity (FC)") +
    scale_fill_manual(values = c("Downregulated" = "#6495ED", 
                                 "NS" = "grey80", 
                                 "Upregulated" = "#F37F80")) +
    scale_alpha_manual(values = c(`TRUE` = 0.4, `FALSE` = 1), guide = "none") +
    theme(aspect.ratio = 1) +
    ggtitle(i)
  
  print(p)
}
```
Conclusion: There's a lot to see here. In some conditions, there's a pretty good overlap between the changing TFs, in other conditions not.

---

## STAT3 changes over time
```{r out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
## Visualize STAT3 fold-changes over time
ggplot(tf_activity_chromVAR_dev %>%
         filter(tf == "STAT3", condition %in% c("mES_LIF_1h", "mES_LIF_3h", "mES_LIF_6h", "mES_LIF")) %>%
         distinct(condition, logFC, sig) %>%
         mutate(condition = factor(condition, levels = c("mES_LIF", "mES_LIF_6h", "mES_LIF_3h", "mES_LIF_1h"))) %>%
         distinct(),
       aes(x = condition, y = logFC, fill = sig)) +
  geom_bar(stat = "identity", width = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +  # Add a vertical line at x = 0
  coord_flip() +  # Flip the plot to make it horizontal
  scale_fill_manual(values = c("Upregulated" = "#F37F80", "NS" = "grey80", "Downregulated" = "#6495ED")) +
  theme(aspect.ratio = 2) +
  ylim(-6, 0)

ggplot(tf_activity_chromVAR_dev %>%
         filter(tf == "STAT3", condition %in% c("mES_LIF_1h", "mES_LIF_3h", "mES_LIF_6h", "mES_LIF")) %>%
         distinct(condition, chromVAR_zscore_dif_mean, sig_chrom) %>%
         mutate(condition = factor(condition, levels = c("mES_LIF", "mES_LIF_6h", "mES_LIF_3h", "mES_LIF_1h"))) %>%
         distinct(),
       aes(x = condition, y = chromVAR_zscore_dif_mean, fill = sig_chrom)) +
  geom_bar(stat = "identity", width = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +  # Add a vertical line at x = 0
  coord_flip() +  # Flip the plot to make it horizontal
  scale_fill_manual(values = c("Upregulated" = "#F37F80", "NS" = "grey80", "Downregulated" = "#6495ED")) +
  theme(aspect.ratio = 2) +
  ylim(-0.27,0)


timepoints <- data.frame(
  condition = c("mES_LIF_1h", "mES_LIF_3h", "mES_LIF_6h", "mES_LIF"),
  timepoint_label = c(1, 3, 6, 24)
)

ggplot(tf_activity_chromVAR_dev %>%
         filter(tf == "STAT3", condition %in% c("mES_LIF_1h", "mES_LIF_3h", "mES_LIF_6h", "mES_LIF")) %>%
         left_join(timepoints) %>%
         distinct(condition, logFC, sig, timepoint_label),
       aes(x = timepoint_label, y = logFC)) +
  geom_line() +
  geom_point(size = 5, aes(color = sig)) +
  scale_color_manual(values = c("Upregulated" = "#F37F80", "NS" = "grey80", "Downregulated" = "#6495ED")) +
  theme(aspect.ratio = .5) +
  ylim(-6, 0) +
  # Set x-axis labels
  scale_x_continuous(breaks = timepoints$timepoint_label, labels = timepoints$timepoint_label) +
  labs(x = "Timepoint (h)", y = "Reporter activity (FC)")

ggplot(tf_activity_chromVAR_dev %>%
         filter(tf == "STAT3", condition %in% c("mES_LIF_1h", "mES_LIF_3h", "mES_LIF_6h", "mES_LIF")) %>%
         left_join(timepoints) %>%
         distinct(condition,  chromVAR_zscore_dif_mean, sig_chrom, timepoint_label),
       aes(x = timepoint_label, y = chromVAR_zscore_dif_mean)) +
  geom_line() +
  geom_point(size = 5, aes(color = sig_chrom)) +
  scale_color_manual(values = c("Upregulated" = "#F37F80", "NS" = "grey80", "Downregulated" = "#6495ED")) +
  theme(aspect.ratio = .5) +
  ylim(-0.27,0) +
  # Set x-axis labels
  scale_x_continuous(breaks = timepoints$timepoint_label, labels = timepoints$timepoint_label) +
  labs(x = "Timepoint (h)", y = "Reporter activity (FC)")

```
Conclusion: TF reporters drop over time, chromVAR signal stays similar.

---

## Effect on target TFs - comparing both methods
```{r out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
## Only target TF - double sided bar plot
target_conditions <- data.frame(condition = c("mES_CH", "mES_FOXA1_OE", "mES_LIF",
                                              "mES_POU5F1_DEG", "mES_SOX2_DEG", "mES_FBS",
                                              "mES_FK", "mES_Heat", "mES_TFCP2L1"),
                                tf = c("TCF7", "FOXA1", "STAT3",
                                       "POU5F1::SOX2", "SOX2", "SRF",
                                       "CREB1", "HSF1", "TFCP2L1"))

tf_activity_chromVAR_dev <- prime_activities_comb %>%
  left_join(chromvar_scores_mouse_added_all %>% 
              mutate(chromVAR_zscore_dif_mean = ifelse(condition == "mES_FOXA1_OE", chromVAR_zscore_dif, chromVAR_zscore_dif_mean)) %>%
              distinct(tf, condition, chromVAR_zscore_dif_mean, p_value_adjusted)) %>%
  distinct(tf, logFC, chromVAR_zscore_dif_mean, condition, sig, p_adjusted, p_value_adjusted) %>%
  mutate(logFC = ave(logFC, tf, condition, FUN = function(x) mean(x, na.rm = T))) %>%
  distinct() %>%
  mutate(sig_chrom = ifelse(p_value_adjusted < 0.05 & chromVAR_zscore_dif_mean > 0, "Upregulated", "NS"),
         sig_chrom = ifelse(p_value_adjusted < 0.05 & chromVAR_zscore_dif_mean < 0, "Downregulated", sig_chrom))


target_plot <- target_conditions %>%
  left_join(tf_activity_chromVAR_dev) %>%
  pivot_longer(cols = c(logFC, chromVAR_zscore_dif_mean),
               values_to = "activity",
               names_to = "method") %>%
  mutate(tf = factor(tf, levels = c("HSF1", "FOXA1", "CREB1",
                                       "SRF", "TFCP2L1", "POU5F1::SOX2",
                                       "SOX2", "STAT3", "TCF7")))

# Create the double-sided bar plot
ggplot(target_plot %>% filter(method == "logFC"), aes(x = tf, y = activity, fill = sig)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +  # Add a vertical line at x = 0
  coord_flip() +  # Flip the plot to make it horizontal
  scale_y_continuous(labels = abs) +  # Display positive labels for both sides
  labs(x = "Category", y = "Value", title = "Double-Sided Bar Plot") +
  scale_fill_manual(values = c("Upregulated" = "#F37F80", "NS" = "grey80", "Downregulated" = "#6495ED")) +
  theme(aspect.ratio = 2) +
  ylim(-6, 6)

ggplot(target_plot %>% filter(method == "chromVAR_zscore_dif_mean"), aes(x = tf, y = activity, fill = sig_chrom)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +  # Add a vertical line at x = 0
  coord_flip() +  # Flip the plot to make it horizontal
  scale_y_continuous(labels = abs) +  # Display positive labels for both sides
  labs(x = "Category", y = "Value", title = "Double-Sided Bar Plot") +
  scale_fill_manual(values = c("Upregulated" = "#F37F80", "NS" = "grey80", "Downregulated" = "#6495ED")) +
  theme(aspect.ratio = 2) +
  ylim(-0.3, 0.3)
```
Conclusion: TF reporters are more sensitive overall.

---

## Number of upregulated & downregulated per condition and method
```{r out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
## Number of upregulated & downregulated per condition and method
up_down_reg <- tf_activity_chromVAR_dev %>%
  distinct(condition, tf, sig, sig_chrom) %>%
  pivot_longer(cols = c(sig, sig_chrom),
               values_to = "sig_comb",
               names_to = "method")

ggplot(up_down_reg, aes(x = condition, fill = sig_comb)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Upregulated" = "#F37F80", "NS" = "grey80", "Downregulated" = "#6495ED")) +
  theme(aspect.ratio = 2) +
  labs(x = "Condition", y = "Proportion of TFs", title = "Proportion of Upregulated and Downregulated TFs") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~method, ncol = 2) +
  theme(legend.position = "bottom")

```

---

## Comparing target TFs rank between chromVAR and primeTF
```{r out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
## Compute ranks
target_conditions <- data.frame(condition = c("mES_CH", "mES_FOXA1_OE", "mES_LIF",
                                              "mES_POU5F1_DEG", "mES_SOX2_DEG", "mES_FBS",
                                              "mES_FK", "mES_Heat", "mES_TFCP2L1"),
                                tf = c("TCF7", "FOXA1", "STAT3",
                                       "POU5F1::SOX2", "SOX2", "SRF",
                                       "CREB1", "HSF1", "TFCP2L1")) %>%
  mutate(target = T)

target_ranks <- tf_activity_chromVAR_dev %>%
  distinct(tf, condition, p_adjusted, p_value_adjusted, chromVAR_zscore_dif_mean, logFC) %>%
  mutate(chromVAR_sig = -log10(p_value_adjusted) * abs(chromVAR_zscore_dif_mean)) %>%
  mutate(primetime_sig = -log10(p_adjusted) * abs(logFC)) %>%
  mutate(chromVAR_sig = ifelse(condition == "mES_FOXA1_OE", chromVAR_zscore_dif_mean, chromVAR_sig)) %>%
  left_join(target_conditions) %>%
  mutate(target = ifelse(is.na(target), F, T)) %>%
  group_by(condition) %>%
  mutate(primetime_rank = rank(-primetime_sig),
         chromVAR_rank = rank(-chromVAR_sig)) %>%
  ungroup() %>%
  distinct(tf, condition, primetime_rank, chromVAR_rank, target) %>%
  filter(target)

ggplot(target_ranks,
       aes(x = chromVAR_rank, y = primetime_rank)) +
  geom_point(size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_text_repel(aes(label = tf), size = 3) +
  labs(x = "chromVAR rank", y = "primeTF rank") +
  xlim(0, 100) +
  ylim(0, 100) +
  theme(aspect.ratio = 1)

# Generate a heatmap with the ranks (using pheatmap)
library(pheatmap)
target_ranks_matrix <- target_ranks %>%
  distinct(tf, chromVAR_rank, primetime_rank) %>%
  column_to_rownames("tf") %>%
  as.matrix() %>%
  t() %>% 
  round(0)

pheatmap(target_ranks_matrix,
         cluster_rows = TRUE,
         border_color = NA,
         cluster_cols = TRUE,
         display_numbers = TRUE,
         fontsize_number = 8,
         fontsize_row = 8,
         fontsize_col = 8,
         main = "Rank comparison between chromVAR and primeTF",
         color = viridis::viridis(50),
         breaks = c(seq(0, 10, length.out = 26), seq(11, 80, length.out = 25)),  # Center around 10
         cellwidth = 20,
         cellheight = 20)
  

```
Conclusion: For some TFs, chromVAR does not pick up well changes in activity.
---

## Compare chromVAR computed using our data with data from de Wit's lab
```{r out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
# Load chromVAR activities computed from de Wit's data
chromvar_scores_teun <- readRDS("/DATA/shared/projects/TF_REPORTERS/ATAC_TF_reporter_comparison/data/chromVAR/dev_scores_teun.rds") %>%
  distinct(tf, chromVAR_zscore_dif_mean, condition) %>%
  mutate(source = "teun") %>%
  mutate(condition = gsub("SOX2.*", "SOX2", condition))

chromvar_scores_mouse_comp <- chromvar_scores_mouse_added_all %>%
  distinct(chromVAR_zscore_dif_mean, tf, condition) %>%
  filter(condition %in% c("mES_POU5F1_DEG", "mES_SOX2_DEG")) %>%
  mutate(condition = ifelse(condition == "mES_POU5F1_DEG", "OCT4", "SOX2")) %>%
  mutate(source = "gen") %>%
  rbind(chromvar_scores_teun) %>%
  spread(source, chromVAR_zscore_dif_mean)

ggplot(chromvar_scores_mouse_comp %>%
         filter(condition == "SOX2"),
       aes(x = gen, y = teun)) +
  geom_hline(yintercept = 0, lty = 1) +
  geom_vline(xintercept = 0, lty = 1) +
  geom_smooth(method = "lm", color = "red") +
  geom_abline(lty = 2) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point(size = 3, shape = 21, fill = "grey") +
  labs(x = "chromVAR deviation z-score dif (gen)", y = "chromVAR deviation z-score dif (teun)") +
  theme(aspect.ratio = 1)
```

---

## Compare chromVAR with logOR of motif accessibility
```{r out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
# Import motif abundance in differential ATAC-seq peaks - computed by Teun on their data
motif_statistics <- readRDS("~/projects/SuRE-TF/analyses/ATAC_degron/motif_statistics_bos_sox2_only.rds") %>%
  filter(time == "24H") %>%
  mutate(motif_name = toupper(motif_name)) %>%
  mutate(motif_name = ifelse(motif_name == "NR1H2::RXRA", "NR1H2", motif_name)) %>%
  mutate(motif_name = ifelse(motif_name == "RARA::RXRA", "RARA:RXRA", motif_name)) %>%
  mutate(motif_name = ifelse(motif_name == "RXRA::VDR", "VDR:RXRA", motif_name)) %>%
  group_by(motif_name) %>%
  mutate(pval_min = min(fdr)) %>%
  filter(fdr == pval_min) %>%
  mutate(ATAC_change = ifelse(contrast == "24H_down", -LOR, LOR)) %>%
  dplyr::select("tf" = motif_name, ATAC_change) %>% 
  distinct() %>%
  left_join(chromvar_scores_teun %>% 
              filter(condition == "SOX2") %>% 
              distinct(chromVAR_zscore_dif_mean, tf)) %>%
  na.omit()

ggplot(motif_statistics,
       aes(x = ATAC_change, y = chromVAR_zscore_dif_mean)) +
  geom_hline(yintercept = 0, lty = 1) +
  geom_vline(xintercept = 0, lty = 1) +
  geom_smooth(method = "lm", color = "red") +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point(size = 3, shape = 21, fill = "grey") +
  theme(aspect.ratio = 1) +
  labs(x = "Log odds ratio (motif accessibility)", y = "chromVAR deviation z-score dif") 
```

---

# Session Info
```{r out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
paste("Run time: ",format(Sys.time()-StartTime))
getwd()
date()
sessionInfo()
```

