---
title: "Comparison of ATAC-seq inferred activity with TF reporter activity"
author: 
  - name: "Max Trauernicht"
    email: "m.trauernicht@nki.nl"
    affiliation: "Netherlands Cancer Institute / University of Twente"
date: '20/02/2025'
output: 
  html_document:
    theme: flatly  # Modern, professional look
    highlight: tango  # Clean code highlighting
    fig_caption: true
    code_folding: show  # Users can expand/collapse code
    df_print: paged  # Paginate long tables instead of cluttering the page
    toc: true
    toc_depth: 2  # Limits TOC depth for clarity
    toc_float: 
      collapsed: true  # Keeps TOC clean, expandable when needed
      smooth_scroll: true  # Smooth scrolling when clicking TOC links
    number_sections: true  # Automatically number sections
    mathjax: "default"  # Enables LaTeX-style equations if needed
---

---

### Aim
Comparing activities between ATAC-seq inferred TF activities and activities measured using my optimized prime TF reporters.

---

## Setup {.tabset}

<!-- little HTML script to do indentation of the table of contents -->
<script>
    $(document).ready(function() {
      $items = $('div#TOC li');
      $items.each(function(idx) {
        num_ul = $(this).parentsUntil('#TOC').length;
        $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
      });

    });
</script>

```{css, echo = FALSE}
div.sourceCode {
  overflow-x: hidden;
}
```

### Libraries 

```{r setup, out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
knitr::opts_chunk$set(echo = TRUE)
StartTime <-Sys.time()

# 8-digit Date tag:
Date <- substr(gsub("-","",Sys.time()),1,8) 
# libraries:
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(tibble)
library(pheatmap)
library(visNetwork)
library(ggbeeswarm)
library(ggforce)
library(viridis)
library(purrr)
library(plyr)
library(igraph)
library(ggraph)
library(cowplot)
library(gridExtra)
library(pROC)
library(tidyr)
library(stringr)
library(ggrastr)
library(readr)
library(ggbiplot)
library(ggh4x)
library(ggiraph)
library(plotly)
library(AMR)
library(umap)
library(ggrepel)
library(chromVAR)
library(Matrix)
library(SummarizedExperiment)
library(BiocParallel)
library(BSgenome.Mmusculus.UCSC.mm10)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ggnewscale)
```

### Functions

```{r out.width= "80%", fig.align= "center"}
### Custom functions
SetFileName <- function(filename, initials) {
  # Set filename with extension and initials to make filename with date integrated.
  filename <- substitute(filename)
  initials <- substitute(initials)
  filename <- paste0(initials, Date, filename)
  filename
}


# Extract p-value from linear model
lmp <- function (modelobject) {
    if (class(modelobject) != "lm") stop("Not an object of class 'lm' ")
    f <- summary(modelobject)$fstatistic
    p <- pf(f[1],f[2],f[3],lower.tail=F)
    attributes(p) <- NULL
    return(p)
}

theme_custom <- theme(
  text = element_text(size = 12),
  
  # Remove panel background
  panel.background = element_blank(),
  
  # Customize grid lines: Keep both major X and Y grid lines, make them thinner
  panel.grid = element_blank(),
  panel.grid.major = element_line(colour = "#e3e1e1", linetype = 1, size = 0.3),
  
  # Ensure axis lines and ticks are visible
  axis.line = element_line(colour = "black"),
  axis.ticks = element_line(colour = "black"),
  
  # Make axis ticks longer
  axis.ticks.length = unit(7, "pt"),  # Adjust the length as needed
  
  # Add an outside border to the plotting area
  panel.border = element_rect(colour = "black", fill = NA, size = .5),
  
  # Ensure facet (strip) outline matches the panel border thickness
  strip.background = element_rect(fill = "grey90", colour = "black", size = .5),
  strip.text = element_text(face = "bold"),
  
  # Adjust plot layout
  plot.title.position = "plot",
  legend.position = "top",
  legend.title = element_blank()
)

theme_set(theme_minimal() + theme_custom)

# Function to force axes to start at zero
force_zero_axes <- function(p) {
  p + 
    scale_x_continuous(expand = c(0, 0), limits = c(0, NA)) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, NA))
}
```

### Loading reporter activities

```{r out.width= "80%", fig.align= "center"}
# Import TF reporter activities (only of the identified prime reporters)
## Import reporter activities
cDNA_df <- read.csv("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/TF_reporter_analyses/data/mt20240724_reporter_activity_filt_combined_complete.csv", header = T) %>%
  mutate(tf = gsub("_.*", "", tf))

# Load RNA-seq data
tf_rna <- read_tsv("/DATA/usr/m.trauernicht/data/RNA_seq/rna_tpm_all_tfs.tsv")

## Import prime reporters
prime_reporters <- read_csv("/DATA/usr/m.trauernicht/projects/SuRE-TF/data/best_reporters.csv") #%>%
  #filter(conf_level > 1)

prime_activities <- cDNA_df %>%
  filter(reporter_id %in% prime_reporters$reporter_id) %>%
  distinct(tf, condition, reporter_activity_dif_sample, replicate, reporter_activity_minP, "ctrl_activity" = reference_activity_minP) %>%
  mutate(reporter_activity_dif_sample = ave(reporter_activity_dif_sample, tf, condition, replicate, FUN = function(x) mean(x, na.rm = T))) %>%
  mutate(reporter_activity_minP = ave(reporter_activity_minP, tf, condition, replicate, FUN = function(x) mean(x, na.rm = T))) %>%
  distinct() %>%
  mutate(replicate = toupper(replicate)) %>%
  filter(condition %in% c("mES_SOX2_DEG", "mES_POU5F1_DEG", "mES_FOXA1_OE", "mES_PD", "mES_CH", "mES_LIF", "U2OS_PMA", "U2OS_Heat", "U2OS_Calcitriol", "MCF7_Hexestrol", "mES_2i_LIF"))

## Import primetime-computed activities of added conditions
base_dir <- "/DATA/usr/v.franceschini/Workspaces/2025_01_13_MAX_CONDITIONS_PRIMETIME/01_OUTPUT"
sub_dirs <- list.dirs(base_dir, recursive = FALSE)
file_paths <- file.path(sub_dirs, "MPRAnalyze", "MPRAnalyze_results.txt")
file_paths <- file_paths[file.exists(file_paths)]
read_and_process <- function(file) {
  data <- read_delim(file) %>%
    select(logFC, tf, p_adjusted, sig) %>%  # Keep only selected columns
    mutate(condition = basename(dirname(dirname(file))))  # Extract unique subdirectory name
  return(data)
}
data_list <- map(file_paths, read_and_process)
primetime_activities <- bind_rows(data_list) %>%
  mutate(condition = gsub("_DMOS", "", condition)) %>%
  mutate(condition = gsub("_DMSO", "", condition)) %>%
  filter(!condition %in% c("ITO60_T_vs_H", "U2OS_DMSO_Calc_v1", "U2OS_DMSO_Calc_v1_5", "U2OS_DMSO_Calc_v2", "U2OS_DMSO_Calc_v2_5"))


## Import primetime-computed activities of all conditions
base_dir <- "/DATA/usr/m.trauernicht/projects/prime_TF_reporters/mES_ATAC_primetime/01_OUTPUT"
sub_dirs <- list.dirs(base_dir, recursive = FALSE)
file_paths <- file.path(sub_dirs, "primetime_results", "primetime_results.txt")
file_paths <- file_paths[file.exists(file_paths)]
read_and_process <- function(file) {
  data <- read_delim(file) %>%
    select(logFC, tf, p_adjusted, sig, "ctrl_activity" = DMSO) %>%  # Keep only selected columns
    mutate(condition = basename(dirname(dirname(file))))  # Extract unique subdirectory name
  return(data)
}
data_list <- map(file_paths, read_and_process)
primetime_activities_all <- bind_rows(data_list) %>%
  filter(str_detect(tf, "RANDOM", negate = T))

primetime_activities_all_1 <- primetime_activities_all %>%
  filter(condition %in% c("mES_CH", "mES_FOXA1", "mES_LIF24", "mES_POU5F1", "mES_SOX2")) %>%
  mutate(condition = ifelse(condition == "mES_LIF24", "mES_LIF", condition)) %>%
  mutate(condition = ifelse(condition == "mES_SOX2", "mES_SOX2_DEG", condition)) %>%
  mutate(condition = ifelse(condition == "mES_POU5F1", "mES_POU5F1_DEG", condition)) %>%
  mutate(condition = ifelse(condition == "mES_FOXA1", "mES_FOXA1_OE", condition)) %>%
  dplyr::select(-logFC, -ctrl_activity)

primetime_activities_all_2 <- primetime_activities_all %>%
  filter(!condition %in% c("mES_CH", "mES_FOXA1", "mES_LIF24", "mES_POU5F1", "mES_SOX2")) %>%
  distinct(tf, condition, logFC, p_adjusted, sig, ctrl_activity)


prime_activities_comb <- prime_activities %>%
  filter(condition %in% c("mES_CH", "mES_FOXA1_OE", "mES_LIF", "mES_POU5F1_DEG", "mES_SOX2_DEG")) %>%
  dplyr::select(tf, condition, "logFC" = reporter_activity_dif_sample, replicate, ctrl_activity) %>%
  left_join(primetime_activities_all_1)

prime_activities_comb <- rbind.fill(prime_activities_comb, primetime_activities_all_2)
```

---

### Load chromVAR data
```{r out.width= "80%", fig.align= "center"}
chromvar_scores_mouse <- readRDS("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/data/chromVAR/dev_scores_mouse.rds")

chromvar_scores_human <- readRDS("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/data/chromVAR/dev_scores_human.rds")

chromvar_scores_mouse_added_mES <- readRDS("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/data/chromVAR/dev_scores_mouse_2025_mES.rds")

chromvar_scores_mouse_added_NPC <- readRDS("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/data/chromVAR/dev_scores_mouse_2025_NPC.rds")

chromvar_scores_mouse_added <- rbind(chromvar_scores_mouse_added_mES, chromvar_scores_mouse_added_NPC)

chromvar_scores_mouse_added_all <- readRDS("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/data/chromVAR/dev_scores_mouse_2025_mouse_all.rds")

```

---

## Correlate new with old mES data
```{r}
chromvar_scores_mouse_old <- chromvar_scores_mouse_added_all %>%
  filter(condition == "mES_2i_LIF") %>%
  dplyr::select(tf, "old_chromVAR" = chromVAR_zscore_mean) %>%
  distinct()

chromvar_scores_mouse_new <- chromvar_scores_mouse_added_all %>%
  filter(condition == "mES_NT") %>%
  dplyr::select(tf, "new_chromVAR" = chromVAR_zscore_mean) %>%
  distinct()

chromvar_scores_mouse_comb <- chromvar_scores_mouse_old %>%
  left_join(chromvar_scores_mouse_new, by = "tf")

ggplot(chromvar_scores_mouse_comb, aes(x = old_chromVAR, y = new_chromVAR)) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(x = "Old chromVAR z-score", y = "New chromVAR z-score") +
  theme(aspect.ratio = 1)
```

---

## Plot chromVAR z-scores upon SOX2 degradation

```{r}
chromvar_scores_mouse_added_sox2 <- chromvar_scores_mouse_added_all %>%
  filter(condition %in% c("mES_SOX2_DEG")) %>%
  distinct(tf, condition, chromVAR_zscore_mean, p_value, ctrl_zscore) %>%
  mutate(ctrl_zscore = ave(ctrl_zscore, tf, FUN = function(x) mean(x))) %>%
  distinct() %>%
  mutate(
    sig = case_when(
      p_value < 0.05 & ctrl_zscore > chromVAR_zscore_mean ~ "Downregulated",
      p_value < 0.05 & ctrl_zscore < chromVAR_zscore_mean ~ "Upregulated", TRUE ~ "NS"), 
    sig = factor(sig, levels = c("NS", "Downregulated", "Upregulated"))) %>%
  mutate(color_axis = ifelse(sig == "Upregulated", "#f37f80", ifelse(sig == "NS", "gray30", "#6495ed"))) %>%
  arrange(desc(ctrl_zscore)) %>%
  mutate(tf = factor(tf, levels = unique(tf))) %>%
  distinct(tf, .keep_all = T) %>%
  arrange(desc(ctrl_zscore)) %>%
  mutate(tf = factor(tf, levels = unique(tf)))


chromvar_scores_mouse_added_sox2 %>%
            ggplot() +
            geom_segment(aes(
                y = ctrl_zscore, 
                yend = chromVAR_zscore_mean,
                x = tf, xend = tf,
                color = sig
            ), size = 1) +
            scale_color_manual(
                values = c("NS" = "grey", "Downregulated" = "#6495ed", "Upregulated" = "#f37f80"),
                name = "Result of\nComparative\nAnalysis"
            ) +
            new_scale_color() +
            geom_point(aes(x = tf, y = ctrl_zscore, color = "C"), size = 3) +
            geom_point(aes(x = tf, y = chromVAR_zscore_mean, color = sig), size = 3) +
            scale_color_manual(values = c('C'= "black", "NS" = "grey", "Downregulated" = "#6495ed", "Upregulated" = "#f37f80")) +
            guides(color = "none") +
            theme_bw() +
            theme(
                axis.text.x = element_text(
                    angle = 90, hjust = 1, vjust = 0.5,
                    color = chromvar_scores_mouse_added_sox2$color_axis
                ),
                # Remove inner lines of y-axis
                panel.grid.major.y = element_blank(),
                panel.grid.minor.y = element_blank(),
                text = element_text(size = 14)
            ) +
            labs(
                y = "Activity (log2(RPM+1))",
                x = "",
                title= paste("mES_SOX2_DEG", "vs.", "mES_SOX2_ctrl")
            )


chromvar_scores_mouse_added_sox2 <- chromvar_scores_mouse_added_all %>%
  filter(condition %in% c("mES_SOX2_DEG")) %>%
  distinct(tf, condition, chromVAR_zscore_dif_mean, p_value) %>%
  mutate(
    sig = case_when(
      p_value < 0.05 & chromVAR_zscore_dif_mean < 0 ~ "Downregulated",
      p_value < 0.05 & chromVAR_zscore_dif_mean > 0 ~ "Upregulated", TRUE ~ "NS"), 
    sig = factor(sig, levels = c("NS", "Downregulated", "Upregulated"))) 

ggplot(chromvar_scores_mouse_added_sox2, aes(x = chromVAR_zscore_dif_mean, y = -log10(p_value), color = sig, alpha = sig, tf=tf)) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") +
  geom_point() +
  # sig will be tab-blue
  scale_color_manual(values = c("NS" = "grey", "Downregulated" = "#6495ed", "Upregulated" = "#f37f80")) +
  scale_alpha_manual(values = c("NS" = 0.4, "Upregulated" = 1, "Downregulated" = 1)) +
  # annotate the significant points
  geom_label_repel(data = . %>% filter(sig != "NS"), size=5, aes(label = tf), box.padding = 0.5) +
  guides(color = "none", alpha = "none") +
  theme(aspect.ratio = 1)

chromvar_scores_mouse_added_sox2 <- chromvar_scores_mouse_added_all %>%
  filter(condition %in% c("mES_SOX2_DEG"), tf == "SOX2") %>%
  distinct(chromVAR_zscore, ctrl_zscore) %>%
  pivot_longer(ctrl_zscore:chromVAR_zscore, names_to = "type", values_to = "zscore")

chromvar_scores_mouse_added_sox2_mean <- chromvar_scores_mouse_added_sox2 %>%
  group_by(type) %>%
  summarise(zscore = mean(zscore))

ggplot() +
  geom_bar(data = chromvar_scores_mouse_added_sox2_mean, aes(x = type, y = zscore), stat = "identity", fill = "grey") +
  geom_point(data = chromvar_scores_mouse_added_sox2, aes(x = type, y = zscore),  width = .1, size = 3, shape = 21, fill = "grey") +
  theme(aspect.ratio = 2.5)



```


## Correlate conditions
```{r}
chromvar_scores_mouse_1 <- chromvar_scores_mouse_added %>%
  filter(condition == "NPC_DMSO") %>%
  dplyr::select(tf, "chromVAR_1" = chromVAR_zscore_mean) %>%
  distinct()

chromvar_scores_mouse_2 <- chromvar_scores_mouse_added %>%
  filter(condition == "mES_NT") %>%
  dplyr::select(tf, "chromVAR_2" = chromVAR_zscore_mean) %>%
  distinct()

chromvar_scores_mouse_comb <- chromvar_scores_mouse_1 %>%
  left_join(chromvar_scores_mouse_2, by = "tf")

ggplot(chromvar_scores_mouse_comb, aes(x = chromVAR_1, y = chromVAR_2)) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(x = "NPC", y = "mESC") +
  theme(aspect.ratio = 1)
```

---

## Correlate inferred activity with measured activity - steady state
```{r}
## Plot inferred TF activity in mESCs
ggplot(chromvar_scores_mouse %>% filter(condition == "mES_2i_LIF"),
       aes(x = reorder(tf, -chromVAR_zscore_mean), y = chromVAR_zscore_mean)) +
  geom_bar(stat = "identity") +
  xlab("TF") +
  ylab("chromVAR z-score") +
  theme(aspect.ratio = 1/4, 
        axis.text.x = element_text(angle = 90, hjust = 1)) 

ggplot(chromvar_scores_mouse %>% filter(condition == "mES_2i_LIF") %>%
         left_join(prime_activities %>% filter(condition == "mES_2i_LIF") %>% distinct(tf, reporter_activity_minP)) %>%
         distinct(tf, chromVAR_zscore_mean, reporter_activity_minP),
       aes(x = chromVAR_zscore_mean, y = log2(reporter_activity_minP))) +
  geom_abline(lty = 2) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  ylim(-4, 13) +
  xlim(-4, 13) +
  geom_point() +
  xlab("chromVAR z-score") +
  ylab("TF reporter activity (log2)") +
  theme(aspect.ratio = 1) 
```

---

## Replicate correlation FC - reporter activity
```{r}
## Correlate chromVAR activities with measured activity
tf_activity_chromVAR_dev <- prime_activities_comb %>%
  distinct(tf, logFC, replicate, condition)

## Correlation between replicates within assays
tf_activity_chromVAR_dev_2 <- tf_activity_chromVAR_dev %>%
  spread(replicate, logFC)

ggplot(tf_activity_chromVAR_dev_2, aes(x = R1, y = R2)) +
  geom_smooth(method = "lm", color = "red") +
  geom_hline(yintercept = 0, lty = 1) +
  geom_vline(xintercept = 0, lty = 1) +
  geom_abline(lty = 2) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point(size = 3, shape = 21, fill = "grey") +
  labs(x = "Reporter activity (FC) R1", y = "Reporter activity (FC) R2") +
  facet_wrap(~condition)

ggplot(tf_activity_chromVAR_dev_2 %>% filter(condition == "mES_SOX2_DEG"), aes(x = R1, y = R2)) +
  geom_smooth(method = "lm", color = "red") +
  geom_hline(yintercept = 0, lty = 1) +
  geom_vline(xintercept = 0, lty = 1) +
  geom_abline(lty = 2) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point(size = 3, shape = 21, fill = "grey") +
  labs(x = "Reporter activity (FC) R1", y = "Reporter activity (FC) R2") +
  facet_wrap(~condition) +
  theme(aspect.ratio = 1)

```

---

## Replicate correlation FC - chromVAR
```{r}
tf_activity_chromVAR_dev_3 <- tf_activity_chromVAR_dev %>%
  dplyr::select(-reporter_activity_dif_sample, -nTPM, -chromVAR_zscore) %>%
  spread(replicate, chromVAR_zscore_dif)

ggplot(tf_activity_chromVAR_dev_3, aes(x = R1, y = R2)) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point() +
  geom_smooth(method = "lm", color = "red") +
  geom_abline(lty = 2) +
  labs(x = "Motif accessibility (FC) R1", y = "Motif accessibility (FC) R2") +
  facet_wrap(~condition)

ggplot(tf_activity_chromVAR_dev_3 %>% filter(condition == "mES_SOX2_DEG"), aes(x = R1, y = R2)) +
  geom_smooth(method = "lm", color = "red") +
  geom_hline(yintercept = 0, lty = 1) +
  geom_vline(xintercept = 0, lty = 1) +
  geom_abline(lty = 2) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point(size = 3, shape = 21, fill = "grey") +
  labs(x = "Motif accessibility (FC) R1", y = "Motif accessibility (FC) R2") +
  theme(aspect.ratio = 1) +
  ylim(-40, 20) +
  xlim(-40, 20) 
```

---

## Correlate chromVAR with primeTF
```{r}
ggplot(tf_activity_chromVAR_dev,
       aes(x = chromVAR_zscore_dif, y = reporter_activity_dif_sample)) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point(aes(size = nTPM)) +
  # add linear regression line and correlation coefficient
  geom_smooth(method = "lm", color = "red") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "chromVAR deviation z-score dif", y = "Reporter activity (FC)") +
  theme(aspect.ratio = 1) +
  facet_grid(condition~replicate)

for (i in unique(tf_activity_chromVAR_dev$condition)) {
  
  p <- ggplot(tf_activity_chromVAR_dev %>%
                filter(condition == i) %>%
                mutate(chromVAR = ave(chromVAR_zscore_dif, tf, condition, FUN = function(x) mean(x, na.rm = T))) %>%
                mutate(tf_activity = ave(reporter_activity_dif_sample, tf, condition, FUN = function(x) mean(x, na.rm = T))) %>%
                distinct(tf, chromVAR, tf_activity, condition, nTPM),
              aes(x = chromVAR, y = tf_activity)) +
    geom_text_repel(aes(label = tf), box.padding = 0.5) +
    geom_point(aes(size = nTPM)) +
    # add linear regression line and correlation coefficient
    geom_smooth(method = "lm", color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(x = "chromVAR deviation z-score dif", y = "Reporter activity (FC)") +
    theme(aspect.ratio = 1) +
    ggtitle(i)
  
  print(p)
}

for (i in unique(tf_activity_chromVAR_dev$condition)) {
  p <- ggplot(tf_activity_chromVAR_dev %>%
                filter(replicate == "R1", condition == i, abs(chromVAR_zscore_dif) > 10),
              aes(x = reorder(tf, -chromVAR_zscore_dif), y = chromVAR_zscore_dif)) +
    geom_bar(stat = "identity") +
    # add linear regression line and correlation coefficient
    labs(y = "chromVAR deviation z-score dif", x = "TF") +
    theme(x.text = element_text(angle = 90, hjust = 1)) +
    theme(aspect.ratio = 1) +
    ggtitle(i)
  
  print(p)
}
```


## Same with only added data - reporter activities calculated by primetime
```{r}
## Correlate chromVAR activities with measured activity
tf_activity_chromVAR_dev <- primetime_activities %>%
  mutate(condition = ifelse(condition == "mESC_LIF1", "mESC_LIF_1h", condition)) %>%
  mutate(condition = ifelse(condition == "mESC_LIF6", "mESC_LIF_6h", condition)) %>%
  mutate(condition = ifelse(condition == "mESC_LIF3", "mESC_LIF_3h", condition)) %>%
  mutate(condition = ifelse(condition == "mESC_HQ1", "mESC_HQ_1h", condition)) %>%
  mutate(condition = ifelse(condition == "mESC_HQ6", "mESC_HQ_6h", condition)) %>%
  mutate(condition = ifelse(condition == "mESC_HQ3", "mESC_HQ_3h", condition)) %>%
  mutate(condition = ifelse(condition == "mESC_PD", "mESC_10PD", condition)) %>%
  left_join(chromvar_scores_mouse_added_all %>% 
              distinct(tf, condition, chromVAR_zscore_dif, chromVAR_zscore, replicate) %>%
              mutate(condition = gsub("mES", "mESC", condition)) %>%
              mutate(condition = ifelse(condition == "NPC_DMSO", "NPC", condition))) %>%
  distinct(tf, logFC, chromVAR_zscore_dif, chromVAR_zscore, replicate, condition, sig) %>%
  na.omit()#%>%
  #filter(nTPM > .1 | tf %in% c("SOX9", "FOXA1"))

ggplot(tf_activity_chromVAR_dev %>%
         distinct(tf, condition, chromVAR_zscore_dif, logFC, sig) %>%
         mutate(chromVAR_zscore_dif = ave(chromVAR_zscore_dif, tf, condition, FUN = mean)) %>%
         distinct() %>%
         filter(!condition %in% c("mESC_PMA", "mESC_vitC", "NPC_cGAMP", "NPC_HQ", "mESC_FK", "mESC_SP1")) %>%
         mutate(condition = as.factor(condition)),
       aes(x = chromVAR_zscore_dif, y = logFC)) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point(aes(color = sig)) +
  # add linear regression line and correlation coefficient
  geom_smooth(method = "lm", color = "black") +
  geom_vline(xintercept = 2.576, linetype = "dashed") +
  geom_vline(xintercept = -2.576, linetype = "dashed") +
  labs(x = "chromVAR deviation z-score dif", y = "Reporter activity (FC)") +
  scale_color_manual(values = c("Downregulated" = "#6495ED", "NS" = "grey80", "Upregulated" = "#F37F80")) +
  #theme_pubr(border = T) +
  theme(aspect.ratio = 1) +
  facet_wrap(~condition)

for (i in unique(tf_activity_chromVAR_dev$condition)) {
  
  print(i)
  
  p <- ggplot(tf_activity_chromVAR_dev %>%
                distinct(tf, condition, chromVAR_zscore_dif, logFC, sig, replicate) %>%
                #mutate(chromVAR_zscore_dif = ave(chromVAR_zscore_dif, tf, condition, FUN = mean)) %>%
                distinct() %>%
                #filter(!condition %in% c("mESC_PMA", "mESC_vitC", "NPC_cGAMP", "NPC_HQ", "mESC_FK", "mESC_SP1")) %>%
                filter(condition == i),
              aes(x = chromVAR_zscore_dif, y = logFC)) +
    geom_text_repel(aes(label = tf), box.padding = 0.5) +
    geom_point(aes(color = sig)) +
    # add linear regression line and correlation coefficient
    geom_smooth(method = "lm", color = "black") +
    geom_vline(xintercept = 2.576, linetype = "dashed") +
    geom_vline(xintercept = -2.576, linetype = "dashed") +
    labs(x = "chromVAR deviation z-score dif", y = "Reporter activity (FC)") +
    scale_color_manual(values = c("Downregulated" = "#6495ED", "NS" = "grey80", "Upregulated" = "#F37F80")) +
    theme(aspect.ratio = 1) +
    ggtitle(i) +
    facet_wrap(~replicate)
  
  print(p)
}

## Visualize STAT3 fold-changes over time
ggplot(tf_activity_chromVAR_dev %>%
         filter(tf == "STAT3", condition %in% c("mESC_LIF_1h", "mESC_LIF_3h", "mESC_LIF_6h")) %>%
         distinct(condition, logFC, chromVAR_zscore_dif) %>%
         pivot_longer(cols = c(logFC, chromVAR_zscore_dif),
                      values_to = "stat3_change",
                      names_to = "method") %>%
         mutate(stat3_change = ave(stat3_change, method, condition, FUN = mean)) %>%
         distinct(),
       aes(x = condition, y = stat3_change)) +
  geom_point(aes(color = method)) +
  #geom_line(aes(group = replicate), color = "grey") +
  labs(x = "Condition", y = "Reporter activity (FC)") +
  theme(aspect.ratio = 1)
```

---


## Updated primetime analysis - all conditions
```{r}
## Barcode counts SOX2 reporters upon SOX2 degradation
## Load data
primetime_sox2_bc <- read_delim("/DATA/usr/m.trauernicht/projects/prime_TF_reporters/mES_ATAC_primetime/01_OUTPUT/mES_SOX2/tmp_primetime/activity/cDNA_counts.txt") %>%
  pivot_longer(cols = c(-tf, -negative_control, -promoter, -barcode),
               names_to = "sample",
               values_to = "reads") %>%
  mutate(rpm = ave(reads, sample, FUN = function(x) (x) / sum(x) *1e6 )) %>%
  filter(tf == "SOX2") %>%
  mutate(condition = gsub("_[1-3]{1}", "", sample))

primetime_pDNA <- read_delim("/DATA/usr/m.trauernicht/projects/prime_TF_reporters/mES_ATAC_primetime/01_OUTPUT/mES_SOX2/tmp_primetime/activity/pDNA_counts.txt") %>%
    mutate(pDNA_rpm = pDNA / sum(pDNA) *1e6 )
  
primetime_sox2_bc <- primetime_sox2_bc %>%
  left_join(primetime_pDNA) %>%
  mutate(activity = rpm / pDNA_rpm) 

primetime_sox2_bc_mean <- primetime_sox2_bc %>%
  mutate(rpm_mean = ave(rpm, condition, FUN = function(x) mean(x, na.rm = T))) %>%
  distinct(rpm_mean, condition)

ggplot() +
  geom_bar(data = primetime_sox2_bc_mean, aes(x = condition, y = rpm_mean), stat = "identity", fill = "grey") +
  geom_jitter(data = primetime_sox2_bc, aes(x = condition, y = rpm), width = .1, size = 3, shape = 21, fill = "grey") +
  theme(aspect.ratio = 2.5)


## Correlate chromVAR activities with measured activity
tf_activity_chromVAR_dev <- prime_activities_comb %>%
  mutate(condition = ifelse(condition == "mES_LIF1", "mES_LIF_1h", condition)) %>%
  mutate(condition = ifelse(condition == "mES_LIF6", "mES_LIF_6h", condition)) %>%
  mutate(condition = ifelse(condition == "mES_LIF3", "mES_LIF_3h", condition)) %>%
  left_join(chromvar_scores_mouse_added_all %>% 
              mutate(chromVAR_zscore_dif = ave(chromVAR_zscore_dif, tf, condition, FUN = function(x) mean(x, na.rm = T))) %>%
              distinct(tf, condition, chromVAR_zscore_dif) %>%
              mutate(condition = ifelse(condition == "NPC_DMSO", "NPC", condition)) %>%
              mutate(condition = ifelse(condition == "mES_2i_LIF", "mES_NPC", condition))) %>%
  distinct(tf, logFC, chromVAR_zscore_dif, condition, sig, ctrl_activity) %>%
  mutate(logFC = ave(logFC, tf, condition, FUN = function(x) mean(x, na.rm = T))) %>%
  distinct()#%>%
#filter(nTPM > .1 | tf %in% c("SOX9", "FOXA1"))

ggplot(tf_activity_chromVAR_dev %>%
         mutate(pert_activity = ctrl_activity + logFC) %>%
         filter(tf == "SOX2", condition %in% c("mES_SOX2_DEG")) %>%
         distinct(pert_activity, ctrl_activity) %>%
         pivot_longer(cols = c(pert_activity, ctrl_activity),
                      values_to = "activity",
                      names_to = "condition"),
       aes(x = condition, y = activity)) +
  geom_bar(stat = "identity", fill = "grey") +
  geom_point(width = .1, size = 3, shape = 21, fill = "grey") +
  theme(aspect.ratio = 2.5)

tf_activity_chromVAR_dev <- prime_activities_comb %>%
  mutate(condition = ifelse(condition == "mES_LIF1", "mES_LIF_1h", condition)) %>%
  mutate(condition = ifelse(condition == "mES_LIF6", "mES_LIF_6h", condition)) %>%
  mutate(condition = ifelse(condition == "mES_LIF3", "mES_LIF_3h", condition)) %>%
  filter(!condition %in% c("mES_PMA", "mES_SP1", "NPC_FK", "NPC_HQ")) %>%
  left_join(chromvar_scores_mouse_added_all %>% 
              distinct(tf, condition, chromVAR_zscore_dif_mean, p_value)) %>%
  distinct(tf, logFC, chromVAR_zscore_dif_mean, condition, sig, p_adjusted, p_value) %>%
  mutate(logFC = ave(logFC, tf, condition, FUN = function(x) mean(x, na.rm = T))) %>%
  distinct() %>%
  mutate(sig_chrom = ifelse(p_value < 0.05 & chromVAR_zscore_dif_mean > 0, "Upregulated", "NS"),
         sig_chrom = ifelse(p_value < 0.05 & chromVAR_zscore_dif_mean < 0, "Downregulated", sig_chrom))


ggplot(tf_activity_chromVAR_dev %>%
         mutate(condition = as.factor(condition),
                x_sig = !is.na(p_value) & p_value < 0.05,
                y_sig = sig != "NS",
                low_sig = !(x_sig | y_sig)),  # insignificant for both metrics
       aes(x = chromVAR_zscore_dif_mean, y = logFC)) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  # Dots with NO outline (x_sig = FALSE)
  geom_point(data = ~ subset(., !x_sig),
             aes(fill = sig, alpha = low_sig),
             shape = 21, size = 2.5, stroke = 0) +
  # Dots with black outline (x_sig = TRUE)
  geom_point(data = ~ subset(., x_sig),
             aes(fill = sig, alpha = low_sig),
             shape = 21, size = 2.5, stroke = 1.2, color = "black") +
  geom_smooth(method = "lm", color = "black") +
  labs(x = "chromVAR deviation z-score dif", y = "Reporter activity (FC)") +
  scale_fill_manual(values = c("Downregulated" = "#6495ED", 
                               "NS" = "grey80", 
                               "Upregulated" = "#F37F80")) +
  scale_alpha_manual(values = c(`TRUE` = 0.4, `FALSE` = 1), guide = "none") +
  theme(aspect.ratio = 1) +
  facet_wrap(~condition)




for (i in unique(tf_activity_chromVAR_dev$condition)) {
  
  print(i)
  
  p <- ggplot(tf_activity_chromVAR_dev %>%
                mutate(condition = as.factor(condition),
                       x_sig = !is.na(p_value) & p_value < 0.05,
                       y_sig = sig != "NS",
                       low_sig = !(x_sig | y_sig)) %>%
                filter(condition == i),  # insignificant for both metrics
              aes(x = chromVAR_zscore_dif_mean, y = logFC)) +
    geom_text_repel(aes(label = tf), box.padding = 0.5) +
    # Dots with NO outline (x_sig = FALSE)
    geom_point(data = ~ subset(., !x_sig),
               aes(fill = sig, alpha = low_sig),
               shape = 21, size = 3, stroke = 0) +
    # Dots with black outline (x_sig = TRUE)
    geom_point(data = ~ subset(., x_sig),
               aes(fill = sig, alpha = low_sig),
               shape = 21, size = 3, stroke = 1.2, color = "black") +
    geom_smooth(method = "lm", color = "black") +
    labs(x = "chromVAR deviation z-score dif", y = "Reporter activity (FC)") +
    scale_fill_manual(values = c("Downregulated" = "#6495ED", 
                                 "NS" = "grey80", 
                                 "Upregulated" = "#F37F80")) +
    scale_alpha_manual(values = c(`TRUE` = 0.4, `FALSE` = 1), guide = "none") +
    theme(aspect.ratio = 1) +
    ggtitle(i)
  
  print(p)
}

## Visualize STAT3 fold-changes over time
ggplot(tf_activity_chromVAR_dev %>%
         filter(tf == "STAT3", condition %in% c("mESC_LIF_1h", "mESC_LIF_3h", "mESC_LIF_6h")) %>%
         distinct(condition, logFC, chromVAR_zscore_dif) %>%
         pivot_longer(cols = c(logFC, chromVAR_zscore_dif),
                      values_to = "stat3_change",
                      names_to = "method") %>%
         mutate(stat3_change = ave(stat3_change, method, condition, FUN = mean)) %>%
         distinct(),
       aes(x = condition, y = stat3_change)) +
  geom_point(aes(color = method)) +
  #geom_line(aes(group = replicate), color = "grey") +
  labs(x = "Condition", y = "Reporter activity (FC)") +
  theme(aspect.ratio = 1)


## Only target TF - double sided bar plot
target_conditions <- data.frame(condition = c("mES_CH", "mES_FOXA1_OE", "mES_LIF",
                                              "mES_POU5F1_DEG", "mES_SOX2_DEG", "mES_FBS",
                                              "mES_FK", "mES_Heat", "mES_TFCP2L1"),
                                tf = c("TCF7", "FOXA1", "STAT3",
                                       "POU5F1::SOX2", "SOX2", "SRF",
                                       "CREB1", "HSF1", "TFCP2L1"))

tf_activity_chromVAR_dev <- prime_activities_comb %>%
  left_join(chromvar_scores_mouse_added_all %>% 
              distinct(tf, condition, chromVAR_zscore_dif_mean, p_value)) %>%
  distinct(tf, logFC, chromVAR_zscore_dif_mean, condition, sig, p_adjusted, p_value) %>%
  mutate(logFC = ave(logFC, tf, condition, FUN = function(x) mean(x, na.rm = T))) %>%
  distinct() %>%
  mutate(sig_chrom = ifelse(p_value < 0.05 & chromVAR_zscore_dif_mean > 0, "Upregulated", "NS"),
         sig_chrom = ifelse(p_value < 0.05 & chromVAR_zscore_dif_mean < 0, "Downregulated", sig_chrom))


target_plot <- target_conditions %>%
  left_join(tf_activity_chromVAR_dev) %>%
  pivot_longer(cols = c(logFC, chromVAR_zscore_dif_mean),
               values_to = "activity",
               names_to = "method")

# Create the double-sided bar plot
ggplot(target_plot %>% filter(method == "logFC"), aes(x = tf, y = activity, fill = sig)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +  # Add a vertical line at x = 0
  coord_flip() +  # Flip the plot to make it horizontal
  scale_y_continuous(labels = abs) +  # Display positive labels for both sides
  labs(x = "Category", y = "Value", title = "Double-Sided Bar Plot") +
  scale_fill_manual(values = c("Upregulated" = "#F37F80", "NS" = "grey80", "Downregulated" = "#6495ED")) +
  theme(aspect.ratio = 2)

ggplot(target_plot %>% filter(method == "chromVAR_zscore_dif_mean"), aes(x = tf, y = activity, fill = sig_chrom)) +
  geom_bar(stat = "identity", width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +  # Add a vertical line at x = 0
  coord_flip() +  # Flip the plot to make it horizontal
  scale_y_continuous(labels = abs) +  # Display positive labels for both sides
  labs(x = "Category", y = "Value", title = "Double-Sided Bar Plot") +
  scale_fill_manual(values = c("Upregulated" = "#F37F80", "NS" = "grey80", "Downregulated" = "#6495ED")) +
  theme(aspect.ratio = 2)
```




## Compare chromVAR computed using our data with data from de Wit's lab
```{r}
# Load chromVAR activities computed from de Wit's data
chromvar_scores_teun <- readRDS("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/data/chromVAR/dev_scores_teun.rds") %>%
  distinct(tf, "chromVAR_zscore_dif_mean" = dif_chromVAR_zscore, condition) %>%
  mutate(source = "teun")

chromvar_scores_mouse_comp <- chromvar_scores_mouse %>%
  distinct(chromVAR_zscore_dif_mean, tf, condition) %>%
  filter(condition %in% c("mES_POU5F1_DEG", "mES_SOX2_DEG")) %>%
  mutate(condition = ifelse(condition == "mES_POU5F1_DEG", "OCT4", "SOX2")) %>%
  mutate(source = "gen") %>%
  rbind(chromvar_scores_teun) %>%
  spread(source, chromVAR_zscore_dif_mean)

ggplot(chromvar_scores_mouse_comp %>%
         filter(condition == "SOX2"),
       aes(x = gen, y = teun)) +
  geom_hline(yintercept = 0, lty = 1) +
  geom_vline(xintercept = 0, lty = 1) +
  geom_smooth(method = "lm", color = "red") +
  geom_abline(lty = 2) +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point(size = 3, shape = 21, fill = "grey") +
  labs(x = "chromVAR deviation z-score dif (gen)", y = "chromVAR deviation z-score dif (teun)") +
  theme(aspect.ratio = 1) +
  ylim(-52, 20) +
  xlim(-52, 20) 
```


## Compare chromVAR with logOR of motif accessibility
```{r}
# Import motif abundance in differential ATAC-seq peaks - computed by Teun on their data
motif_statistics <- readRDS("/DATA/usr/m.trauernicht/projects/SuRE-TF/analyses/ATAC_degron/motif_statistics_bos_sox2_only.rds") %>%
  filter(time == "24H") %>%
  mutate(motif_name = toupper(motif_name)) %>%
  mutate(motif_name = ifelse(motif_name == "NR1H2::RXRA", "NR1H2", motif_name)) %>%
  mutate(motif_name = ifelse(motif_name == "RARA::RXRA", "RARA:RXRA", motif_name)) %>%
  mutate(motif_name = ifelse(motif_name == "RXRA::VDR", "VDR:RXRA", motif_name)) %>%
  group_by(motif_name) %>%
  mutate(pval_min = min(fdr)) %>%
  filter(fdr == pval_min) %>%
  mutate(ATAC_change = ifelse(contrast == "24H_down", -LOR, LOR)) %>%
  dplyr::select("tf" = motif_name, ATAC_change) %>% 
  distinct() %>%
  left_join(chromvar_scores_teun %>% 
              filter(condition == "SOX2") %>% 
              distinct(chromVAR_zscore_dif_mean, tf)) %>%
  na.omit()

ggplot(motif_statistics,
       aes(x = ATAC_change, y = chromVAR_zscore_dif_mean)) +
  geom_hline(yintercept = 0, lty = 1) +
  geom_vline(xintercept = 0, lty = 1) +
  geom_smooth(method = "lm", color = "red") +
  geom_text_repel(aes(label = tf), box.padding = 0.5) +
  geom_point(size = 3, shape = 21, fill = "grey") +
  theme(aspect.ratio = 1) +
  labs(x = "Log odds ratio (motif accessibility)", y = "chromVAR deviation z-score dif") 
```

---


## Compare TF activity to RNA-seq change (DESeq2)
```{r}
# Load DESeq2 results
tf_deseq <- readRDS("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/RNA_seq_analyses/deseq_changed_LIF.rds") %>%
  mutate(tf = ifelse(tf == "ARNT", "AHR::ARNT", tf)) %>%
  mutate(tf = ifelse(tf == "FOS", "FOS::JUN", tf)) %>%
  mutate(tf = ifelse(tf == "NFE2", "MAF::NFE2", tf)) %>%
  mutate(tf = ifelse(tf == "NR1H4", "NR1H4::RXRA", tf)) %>%
  mutate(tf = ifelse(tf == "NR4A2", "NR4A2::RXRA", tf)) %>%
  mutate(tf = ifelse(tf == "PPARA", "PPARA::RXRA", tf)) %>%
  mutate(tf = ifelse(tf == "RARA", "RARA:RXRA", tf)) %>%
  mutate(tf = ifelse(tf == "STAT2", "STAT1::STAT2", tf)) %>%
  mutate(tf = ifelse(tf == "TRP53", "TP53", tf)) %>%
  mutate(tf = ifelse(tf == "VDR", "VDR::RXRA", tf))

tf_deseq_pou <- tf_deseq %>%
  filter(tf == "POU5F1") %>%
  mutate(tf = ifelse(tf == "POU5F1", "POU5F1::SOX2", tf))

tf_deseq <- rbind(tf_deseq, tf_deseq_pou)

prime_activities_deseq <- prime_activities %>%
  filter(condition == "mES_LIF") %>%
  distinct(tf, reporter_activity_dif_sample) %>%
  mutate(reporter_activity_dif_sample = ave(reporter_activity_dif_sample, tf, FUN = mean)) %>%
  distinct() %>%
  left_join(tf_deseq)
  
ggplot(prime_activities_deseq,
       aes(x = reporter_activity_dif_sample, y = lfc)) +
  geom_point(aes(size = basal_exp)) +
  geom_label_repel(aes(label = tf), box.padding = 0.5, point.padding = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  geom_abline(lty = 2) +
  ylim(-4,2) +
  xlim(-4,2) +
  labs(x = "TF activity change (Reporters)", y = "TF abundance log fold change (RNA-seq)") +
  theme_pubr(legend = "none") +
  labs_pubr()

```

---

## Compare reporter activity to VIPER-inferred TF activity
```{r}
# Load VIPER results
viper_activities <- readRDS("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/RNA_seq_analyses/tf_activities_viper.rds")  %>%
  mutate(tf = ifelse(tf == "ARNT", "AHR::ARNT", tf)) %>%
  mutate(tf = ifelse(tf == "FOS", "FOS::JUN", tf)) %>%
  mutate(tf = ifelse(tf == "NFE2", "MAF::NFE2", tf)) %>%
  mutate(tf = ifelse(tf == "NR1H4", "NR1H4::RXRA", tf)) %>%
  mutate(tf = ifelse(tf == "NR4A2", "NR4A2::RXRA", tf)) %>%
  mutate(tf = ifelse(tf == "PPARA", "PPARA::RXRA", tf)) %>%
  mutate(tf = ifelse(tf == "RARA", "RARA:RXRA", tf)) %>%
  mutate(tf = ifelse(tf == "STAT2", "STAT1::STAT2", tf)) %>%
  mutate(tf = ifelse(tf == "TRP53", "TP53", tf)) %>%
  mutate(tf = ifelse(tf == "VDR", "VDR::RXRA", tf))

viper_activities_pou <- viper_activities %>%
  filter(tf == "POU5F1") %>%
  mutate(tf = ifelse(tf == "POU5F1", "POU5F1::SOX2", tf))

viper_activities_dif <- rbind(viper_activities, viper_activities_pou) %>%
  mutate(condition = gsub("_r[1-3]{1}$", "", sample)) %>%
  mutate(viper_activity = ave(viper_activity, tf, condition, FUN = mean)) %>%
  distinct(viper_activity, tf, condition) %>%
  mutate(condition = ifelse(condition == "mES_2i", "mES_LIF", condition)) %>%
  spread(condition, viper_activity) %>%
  mutate(viper_dif = mES_LIF - mES_2i_LIF) %>%
  pivot_longer(cols = c("mES_LIF", "mES_2i_LIF"), names_to = "condition", 
               values_to = "viper_activity") %>%
  mutate(viper_dif = ifelse(condition == "mES_2i_LIF", NA, viper_dif))

prime_activities_viper <- prime_activities %>%
  filter(condition %in% c("mES_LIF", "mES_2i_LIF")) %>%
  distinct(tf, condition, reporter_activity_dif_sample, reporter_activity_minP) %>%
  mutate(reporter_activity_dif_sample = ave(reporter_activity_dif_sample, tf, condition, FUN = mean)) %>%
  mutate(reporter_activity_minP = ave(reporter_activity_minP, tf, condition, FUN = mean)) %>%
  distinct() %>%
  left_join(viper_activities_dif)

ggplot(prime_activities_viper %>%
         filter(condition == "mES_LIF"),
       aes(x = viper_dif, y = reporter_activity_dif_sample)) +
  geom_point() +
  geom_label_repel(aes(label = tf), box.padding = 0.5, point.padding = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  geom_abline(lty = 2) +
  labs(x = "TF activity change (VIPER)", y = "TF activity change (reporters)") +
  theme_pubr(legend = "none") +
  labs_pubr()

ggplot(prime_activities_viper,
       aes(x = viper_activity, y = log2(reporter_activity_minP))) +
  geom_point() +
  geom_label_repel(aes(label = tf), box.padding = 0.5, point.padding = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  geom_abline(lty = 2) +
  labs(x = "TF activity (VIPER)", y = "TF activity (reporters)") +
  theme_pubr(legend = "none") +
  
  facet_wrap(~condition)
```

---

## Compare reporter activity to ISMARA-inferred TF activity
```{r}
# Load VIPER results
ismara_activities <- readRDS("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/RNA_seq_analyses/ismara_activities.rds")  %>%
  mutate(tf = ifelse(tf == "ARNT", "AHR::ARNT", tf)) %>%
  mutate(tf = ifelse(tf == "FOS", "FOS::JUN", tf)) %>%
  mutate(tf = ifelse(tf == "NFE2", "MAF::NFE2", tf)) %>%
  mutate(tf = ifelse(tf == "NR1H4", "NR1H4::RXRA", tf)) %>%
  mutate(tf = ifelse(tf == "NR4A2", "NR4A2::RXRA", tf)) %>%
  mutate(tf = ifelse(tf == "PPARA", "PPARA::RXRA", tf)) %>%
  mutate(tf = ifelse(tf == "RARA", "RARA:RXRA", tf)) %>%
  mutate(tf = ifelse(tf == "STAT2", "STAT1::STAT2", tf)) %>%
  mutate(tf = ifelse(tf == "TRP53", "TP53", tf)) %>%
  mutate(tf = ifelse(tf == "VDR", "VDR::RXRA", tf)) %>%
  mutate(tf = ifelse(tf == "TCF7_TCF7L2", "TCF7", tf)) %>%
  mutate(tf = ifelse(tf == "STAT4_STAT3_STAT5B", "STAT3", tf))

ismara_activities_pou <- ismara_activities %>%
  filter(tf == "POU5F1") %>%
  mutate(tf = ifelse(tf == "POU5F1", "POU5F1::SOX2", tf))

ismara_activities <- rbind(ismara_activities, ismara_activities_pou) %>%
  mutate(condition = ifelse(condition == "mES_2i", "mES_LIF", condition))

prime_activities_ismara <- prime_activities %>%
  filter(condition %in% c("mES_LIF", "mES_2i_LIF")) %>%
  distinct(tf, condition, reporter_activity_dif_sample, reporter_activity_minP) %>%
  mutate(reporter_activity_dif_sample = ave(reporter_activity_dif_sample, tf, condition, FUN = mean)) %>%
  mutate(reporter_activity_minP = ave(reporter_activity_minP, tf, condition, FUN = mean)) %>%
  distinct() %>%
  left_join(ismara_activities)

ggplot(prime_activities_ismara %>%
         filter(condition == "mES_LIF") %>%
         distinct(tf, ismara_activity_dif, reporter_activity_dif_sample, replicate),
       aes(x = ismara_activity_dif, y = reporter_activity_dif_sample)) +
  geom_point() +
  geom_label_repel(aes(label = tf), box.padding = 0.5, point.padding = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  geom_abline(lty = 2) +
  labs(x = "TF activity change (VIPER)", y = "TF activity change (reporters)") +
  theme_pubr(legend = "none") +
  
  facet_wrap(~replicate)

# ggplot(prime_activities_ismara,
#        aes(x = viper_activity, y = log2(reporter_activity_minP))) +
#   geom_point() +
#   geom_label_repel(aes(label = tf), box.padding = 0.5, point.padding = 0.5) +
#   geom_smooth(method = "lm", color = "red") +
#   geom_abline(lty = 2) +
#   labs(x = "TF activity (VIPER)", y = "TF activity (reporters)") +
#   theme_pubr(legend = "none") +
#   
#   facet_wrap(~condition)
```





# Session Info
```{r}
paste("Run time: ",format(Sys.time()-StartTime))
getwd()
date()
sessionInfo()
```

