---
title: "ChromVAR activity computation from ATAC-seq data"
author: 
  - name: "Max Trauernicht"
    email: "m.trauernicht@nki.nl"
    affiliation: "Netherlands Cancer Institute - van Steensel lab"

date: '`r format(Sys.time(), "%d/%m/%Y")`'
output:
  html_document:
    theme: united
    highlight: pygments
    fig_caption: yes
    code_folding: hide
    df_print: kable
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
---

---

### Aim
ATAC-seq was performed in mESCs and mNPCs under various conditions altering TF levels or signaling pathways. Here, we compute chromVAR deviations for a selected set of TF motifs to assess changes in motif activity across conditions.

---

## Setup {.tabset}

<!-- little HTML script to do indentation of the table of contents -->
<script>
    $(document).ready(function() {
      $items = $('div#TOC li');
      $items.each(function(idx) {
        num_ul = $(this).parentsUntil('#TOC').length;
        $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
      });

    });
</script>

```{css, echo = FALSE}
div.sourceCode {
  overflow-x: hidden;
}
```

### Libraries 

```{r setup, out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
knitr::opts_chunk$set(echo = TRUE)
StartTime <-Sys.time()

# 8-digit Date tag:
Date <- substr(gsub("-","",Sys.time()),1,8) 
# libraries:
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(tibble)
library(pheatmap)
library(visNetwork)
library(ggbeeswarm)
library(ggforce)
library(viridis)
library(plyr)
library(igraph)
library(ggraph)
library(cowplot)
library(gridExtra)
library(pROC)
library(tidyr)
library(stringr)
library(ggrastr)
library(readr)
library(ggbiplot)
library(ggh4x)
library(ggiraph)
library(plotly)
library(AMR)
library(umap)
library(ggrepel)
library(chromVAR)
library(Matrix)
library(SummarizedExperiment)
library(BiocParallel)
library(BSgenome.Mmusculus.UCSC.mm10)
library(BSgenome.Hsapiens.UCSC.hg38)
library(universalmotif)
library(motifmatchr)
library(limma)
```

### Functions

```{r out.width= "80%", fig.align= "center"}
### Custom functions
SetFileName <- function(filename, initials) {
  # Set filename with extension and initials to make filename with date integrated.
  filename <- substitute(filename)
  initials <- substitute(initials)
  filename <- paste0(initials, Date, filename)
  filename
}


# Extract p-value from linear model
lmp <- function (modelobject) {
    if (class(modelobject) != "lm") stop("Not an object of class 'lm' ")
    f <- summary(modelobject)$fstatistic
    p <- pf(f[1],f[2],f[3],lower.tail=F)
    attributes(p) <- NULL
    return(p)
}
```

### Loading ATAC-seq data - mouse cells

```{r out.width= "80%", fig.align= "center"}
# Import peak count information for ChromVAR
peak_count <- readRDS("~/projects/ATAC_TF_reporter_comparison/ATAC_seq_analyses/rds/experiment_7784_20250206.rds")
assayNames(peak_count) <- "counts"

# Filter samples
#peak_count <- peak_count[, peak_count@colData$celltype == "mES"] 
peak_count <- peak_count[, peak_count@colData$treatment %in% c("2i_LIF", "NT", "CH", "POU5F1_DEG", "POU5F1_ctrl", "SOX2_DEG",
                                                               "SOX2_ctrl", "FBS", "FK", "Heat", "LIF", "LIF_1h", "LIF_3h",
                                                               "LIF_6h", "TFCP2L1", "FOXA1_OE", "FOXA1_ctrl", "DMSO")] 
peak_count <- peak_count[, peak_count@colData$basename != "1_mES_r3_TAAGGCGA-AGGCTTAG_S1"] 

min_samples <- 1  # Adjust as needed
valid_groups <- names(which(table(peak_count@colData$treatment) >= min_samples))
peak_count <- peak_count[, peak_count@colData$treatment %in% valid_groups]


# Add GC bias
peak_count <- addGCBias(peak_count, genome = BSgenome.Mmusculus.UCSC.mm10)

## Filter peaks - each peak must have at least one count in one condition
peak_count <- filterPeaks(sort(peak_count), non_overlapping = TRUE)

## Extract metadata
metadata <- data.frame(peak_count@colData) %>%
  rownames_to_column("sample") %>%
  distinct(sample, run, celltype, treatment, replicate)


# Import motif information
seq_logos <- readRDS("/DATA/shared/projects/TF_REPORTERS/SuRE-TF/data/library_design/lambert_2018/seq_logos.rds")
motifs_selected <- readRDS("/DATA/shared/projects/TF_REPORTERS/SuRE-TF/data/library_design/lambert_2018/motifs_selected.rds") %>%
  mutate(tf = toupper(tf)) %>%
  mutate(tf = ifelse(tf == "AHR:ARNT", "AHR::ARNT", tf)) %>%
  mutate(tf = ifelse(tf == "RAR:RXR", "RARA:RXRA", tf)) %>%
  rbind(data.frame(tf = c("STAT1", "STAT4", "ATF2", "ATF4", "ATF6", "CLOCK", 
                          "TFEB", "MYC", "CEBPA", "HIF1A", "GATA3", "IRF1", "STAT6", "SREBF1"),
                   motif_id = c("M06794_1.94d.txt", "M06797_1.94d.txt", "M08481_1.94d.txt", "M08489_1.94d.txt", "M09435_1.94d.txt", "M05596_1.94d.txt",
                                "M06206_1.94d.txt", "M06652_1.94d.txt", "M00400_1.94d.txt", "MA1106.1.jaspar", "M05753_1.94d.txt",
                                "M06920_1.94d.txt", "M06798_1.94d.txt", "M09969_1.94d.txt")))
  
seq_logos_pwm <- list()
for (i in names(seq_logos)) {
  if (i %in% motifs_selected$motif_id) {
  seq_logos_pwm[[i]] <- universalmotif::convert_motifs(seq_logos[[i]], class = "TFBSTools-PWMatrix")
  seq_logos_pwm[[i]]@name <- i
  seq_logos_pwm[[i]]@ID <- motifs_selected$tf[motifs_selected$motif_id == i]
  seq_logos_pwm[[i]]@strand <- "+-"
  }
}
# 
# seq_logos_pwm <- list()
# for (i in names(seq_logos)) {
#   seq_logos_pwm[[i]] <- universalmotif::convert_motifs(seq_logos[[i]], class = "TFBSTools-PWMatrix")
#   name_tf <- all_motifs$tf[all_motifs$motif_id == i]
#   seq_logos_pwm[[i]]@name <- name_tf
#   seq_logos_pwm[[i]]@strand <- "+-"
# }
# 

seq_logos_pwm_all <- do.call(TFBSTools::PWMatrixList, seq_logos_pwm)

# Export PWMatrixList


motif_ix <- matchMotifs(seq_logos_pwm_all, peak_count, 
                        genome = BSgenome.Mmusculus.UCSC.mm10)
```

---

### Run chromVAR
```{r out.width= "80%", fig.align= "center"}
# Compute deviations
bg <- getBackgroundPeaks(object = peak_count, niterations = 2000)
dev <- computeDeviations(object = peak_count, annotations = motif_ix
                         ,background_peaks = bg
                         )

dev_scores <- deviations(dev) %>%
  as.data.frame() %>%
  rownames_to_column(var = "motif_id") %>%
  left_join(motifs_selected) %>%
  pivot_longer(cols = -c(motif_id, tf), names_to = "sample", values_to = "chromVAR_zscore") %>%
  # mutate(replicate = gsub(".*(R[1-3]{1}$)", "\\1", sample)) %>%
  mutate(condition = gsub("_R[1-4]{1}$", "", sample)) %>%
  left_join(metadata, by = "sample") %>%
  mutate(sample = gsub("\\.1$", "", sample)) %>%
  mutate(sample = paste(sample, run, sep = "_")) %>%
  mutate(chromVAR_zscore_mean = ave(chromVAR_zscore, motif_id, condition, FUN = mean)) %>%
  distinct(motif_id, tf, condition, run, replicate, chromVAR_zscore_mean, chromVAR_zscore)

dev_scores_ctrl <- dev_scores %>%
  filter(condition %in% c("mES_2i_LIF", "mES_FOXA1_ctrl", "mES_POU5F1_ctrl", "mES_SOX2_ctrl", "mES_NT")) %>%
  mutate(replicate = as.character(replicate)) %>%
  dplyr::select("ref_condition" = condition, "ctrl_zscore" = chromVAR_zscore, tf, replicate, run) %>%
  # mutate(replicate = ifelse(ref_condition == "NPC_DMSO", "R1", replicate)) %>%
  # mutate(run = ifelse(ref_condition == "NPC_DMSO", 8010, run)) %>%
  mutate(replicate = as.factor(replicate))


condition_match <- data.frame(condition = c("mES_LIF", "mES_CH", "mES_FOXA1_OE", "mES_POU5F1_DEG", "mES_SOX2_DEG", "mES_TFCP2L1",
                                            "mES_LIF_1h", "mES_LIF_3h", "mES_LIF_6h", "mES_FBS",
                                            "mES_FK", "mES_Heat", "NPC_DMSO"),
                              ref_condition = c("mES_2i_LIF",  "mES_2i_LIF", "mES_FOXA1_ctrl", "mES_POU5F1_ctrl", "mES_SOX2_ctrl", "mES_NT",
                                                "mES_NT", "mES_NT", "mES_NT", "mES_NT",
                                                "mES_NT", "mES_NT", "mES_NT"))

dev_scores_dif <- dev_scores %>%
  mutate(replicate = as.character(replicate), replicate = ifelse(condition == "NPC_DMSO" & replicate == "R4", "R2", replicate)) %>%
  left_join(condition_match) %>%
  left_join(dev_scores_ctrl) %>%
  mutate(chromVAR_zscore_dif = chromVAR_zscore - ctrl_zscore) %>%
  mutate(chromVAR_zscore_dif_mean = ave(chromVAR_zscore_dif, tf, condition, FUN = function (x) mean(x, na.rm = T))) %>%
  filter(condition %in% c("mES_CH", "mES_LIF", "mES_POU5F1_DEG", "mES_SOX2_DEG", "mES_FBS", "mES_FK", "mES_Heat",
                          "mES_LIF_1h", "mES_LIF_3h", "mES_LIF_6h", "mES_TFCP2L1", "mES_FOXA1_OE", "NPC_DMSO"))

devMat <- assays(dev)$deviations

# Estimate the fold changes and standard errors by fitting a linear model for each gene

condition <- factor(colData(dev)$treatment)
design <- model.matrix(~ 0 + condition)
colnames(design) <- make.names(levels(condition))

# Fit model once to all data
fit <- lmFit(devMat, design)

# List of named contrasts — use clean condition names (no quotes)
contrast_list <- list(
  `CH_vs_2i_LIF` = c("CH", "2i_LIF"),
  `LIF_vs_2i_LIF` = c("LIF", "NT"),
  `Heat_vs_NT` = c("Heat", "NT"),
  `POU5F1_DEG_vs_ctrl` = c("POU5F1_DEG", "POU5F1_ctrl"),
  `SOX2_DEG_vs_ctrl` = c("SOX2_DEG", "SOX2_ctrl"),
  `FBS_vs_NT` = c("FBS", "NT"),
  `FK_vs_NT` = c("FK", "NT"),
  `LIF_1h_vs_NT` = c("LIF_1h", "NT"),
  `LIF_3h_vs_NT` = c("LIF_3h", "NT"),
  `LIF_6h_vs_NT` = c("LIF_6h", "NT"),
  `TFCP2L1_vs_NT` = c("TFCP2L1", "NT"),
  `FOXA1_OE_vs_ctrl` = c("FOXA1_OE", "FOXA1_ctrl"),
  `NPC_mES` = c("DMSO", "NT")
)

# Run contrasts in a loop
results_list <- list()

for (i in 1:13) {
  
  print(i)
  
  groups <- contrast_list[[i]]
  contrast_names <- make.names(groups)

  # Skip if contrast levels don't exist in design
  if (!all(contrast_names %in% colnames(design))) {
    warning(paste("Skipping contrast", name, "— missing levels"))
    return(NULL)
  }

  # Create contrast matrix for this comparison
  contrast_matrix <- makeContrasts(
    contr = paste0(contrast_names[1], " - ", contrast_names[2]),
    levels = design
  )

  # Fit and apply eBayes for this contrast
  fit_contrast <- contrasts.fit(fit, contrast_matrix)
  fit_contrast <- eBayes(fit_contrast)

  # Extract table of results
  top <- topTable(fit_contrast, number = Inf)
  top$motif_id <- rownames(top)
  top$group1 <- groups[1]
  top$group2 <- groups[2]

  results_list[[i]] <- top
}

# Combine into one dataframe
combined_top_table <- bind_rows(results_list) %>%
  left_join(motifs_selected) %>%
  dplyr::select(tf, "chromVAR_zscore_dif_mean" = logFC, "p_value_adjusted" = adj.P.Val, "condition" = group1) %>%
  mutate(condition = paste("mES_", condition, sep = "")) %>%
  mutate(condition = ifelse(condition == "mES_DMSO", "NPC_DMSO", condition))


dev_scores_dif <- dev_scores_dif %>%
  dplyr::select(-chromVAR_zscore_dif_mean) %>%
  left_join(combined_top_table)

saveRDS(dev_scores_dif, "/DATA/shared/projects/TF_REPORTERS/ATAC_TF_reporter_comparison/data/chromVAR/dev_scores_mouse_2025b_mouse_all_npc.rds")
```

---



## ChromVAR on all motifs - check how much the selected TFs explain
```{r out.width= "80%", fig.align= "center"}
# Import peak count information for ChromVAR
peak_count <- readRDS("~/projects/ATAC_TF_reporter_comparison/ATAC_seq_analyses/rds/experiment_7784_20250206.rds")
assayNames(peak_count) <- "counts"

# Filter samples
#peak_count <- peak_count[, peak_count@colData$celltype == "mES"] 
peak_count <- peak_count[, peak_count@colData$treatment %in% c("NT", "DMSO")] 
peak_count <- peak_count[, peak_count@colData$basename != "1_mES_r3_TAAGGCGA-AGGCTTAG_S1"] 

min_samples <- 1  # Adjust as needed
valid_groups <- names(which(table(peak_count@colData$treatment) >= min_samples))
peak_count <- peak_count[, peak_count@colData$treatment %in% valid_groups]


# Add GC bias
peak_count <- addGCBias(peak_count, genome = BSgenome.Mmusculus.UCSC.mm10)

## Filter peaks - each peak must have at least one count in one condition
peak_count <- filterPeaks(sort(peak_count), non_overlapping = TRUE)

## Extract metadata
metadata <- data.frame(peak_count@colData) %>%
  rownames_to_column("sample") %>%
  distinct(sample, run, celltype, treatment, replicate)

seq_logos_pwm <- list()
for (i in names(seq_logos)) {
  seq_logos_pwm[[i]] <- universalmotif::convert_motifs(seq_logos[[i]], class = "TFBSTools-PWMatrix")
  seq_logos_pwm[[i]]@name <- i
  seq_logos_pwm[[i]]@strand <- "+-"
}

seq_logos_pwm_all <- do.call(TFBSTools::PWMatrixList, seq_logos_pwm)

motif_ix <- matchMotifs(seq_logos_pwm_all, peak_count, 
                        genome = BSgenome.Mmusculus.UCSC.mm10)

# Compute deviations
bg <- getBackgroundPeaks(object = peak_count, niterations = 2000)
dev <- computeDeviations(object = peak_count, annotations = motif_ix
                         ,background_peaks = bg
                         )

dev_scores <- deviations(dev) %>%
  as.data.frame() %>%
  rownames_to_column(var = "motif_id") %>%
  mutate(selected = ifelse(motif_id %in% motifs_selected$motif_id, "selected", "all")) %>%
  pivot_longer(cols = -c(motif_id, selected), names_to = "sample", values_to = "chromVAR_zscore") %>%
  # mutate(replicate = gsub(".*(R[1-3]{1}$)", "\\1", sample)) %>%
  mutate(condition = gsub("_R[1-4]{1}$", "", sample)) %>%
  left_join(metadata, by = "sample") %>%
  mutate(sample = gsub("\\.1$", "", sample)) %>%
  mutate(sample = paste(sample, run, sep = "_")) %>%
  mutate(chromVAR_zscore_mean = ave(chromVAR_zscore, motif_id, condition, FUN = mean)) %>%
  distinct(motif_id, selected, condition, run, replicate, chromVAR_zscore_mean, chromVAR_zscore)


dev_scores_ctrl <- dev_scores %>%
  filter(condition %in% c("mES_NT")) %>%
  mutate(replicate = as.character(replicate)) %>%
  dplyr::select("ref_condition" = condition, "ctrl_zscore" = chromVAR_zscore, replicate, run) %>%
  # mutate(replicate = ifelse(ref_condition == "NPC_DMSO", "R1", replicate)) %>%
  # mutate(run = ifelse(ref_condition == "NPC_DMSO", 8010, run)) %>%
  mutate(replicate = as.factor(replicate))


condition_match <- data.frame(condition = c("NPC_DMSO"),
                              ref_condition = c("mES_NT"))

dev_scores_dif <- dev_scores %>%
  mutate(replicate = as.character(replicate), replicate = ifelse(condition == "NPC_DMSO" & replicate == "R4", "R2", replicate)) %>%
  left_join(condition_match) %>%
  left_join(dev_scores_ctrl) %>%
  mutate(chromVAR_zscore_dif = chromVAR_zscore - ctrl_zscore) %>%
  mutate(chromVAR_zscore_dif_mean = ave(chromVAR_zscore_dif, motif_id, condition, FUN = function (x) mean(x, na.rm = T))) %>%
  filter(condition %in% c("NPC_DMSO"))

dev_scores_dif_filt <- dev_scores_dif %>%
  distinct(motif_id, selected, chromVAR_zscore_dif_mean)

ggplot(dev_scores_dif_filt,
       aes(x = selected, y = abs(chromVAR_zscore_dif_mean), color = selected)) +
  geom_quasirandom() 

ggplot(dev_scores_dif_filt,
       aes(x = abs(chromVAR_zscore_dif_mean), color = selected, fill = selected)) +
  geom_density(alpha = .2)

# Statistic comparison of densities
wilcox.test(dev_scores_dif_filt$chromVAR_zscore_dif_mean[dev_scores_dif_filt$selected == "selected"],
            dev_scores_dif_filt$chromVAR_zscore_dif_mean[dev_scores_dif_filt$selected == "all"])

dev_scores_dif_filt2 <- dev_scores_dif_filt %>%
  mutate(sum_zscore = ave(chromVAR_zscore_dif_mean, selected, FUN = function (x) sum(abs(x), na.rm = T))) %>%
  distinct(selected, sum_zscore)



```






## ChromVAR on Teun's data

### Loading data

```{r out.width= "80%", fig.align= "center"}
# Import peak count information for ChromVAR
peak_count <- readRDS("~/projects/SuRE-TF/analyses/ATAC_degron/timeseries_SummarizedExperiment.rds")
assayNames(peak_count) <- "counts"

# Add GC bias
peak_count <- addGCBias(peak_count, genome = BSgenome.Mmusculus.UCSC.mm10)

## Filter peaks - each peak must have at least one count in one condition
peak_count <- filterPeaks(sort(peak_count), non_overlapping = TRUE)

## Extract metadata
metadata <- data.frame(peak_count@colData) %>%
  rownames_to_column("sample") %>%
  distinct(sample, run, celltype, treatment, replicate)


seq_logos_pwm_all <- do.call(TFBSTools::PWMatrixList, seq_logos_pwm)

motif_ix <- matchMotifs(seq_logos_pwm_all, peak_count, 
                        genome = BSgenome.Mmusculus.UCSC.mm10)
```

---

### Run chromVAR
```{r out.width= "80%", fig.align= "center"}
# Compute deviations
#bg <- getBackgroundPeaks(object = sox2_peak_count_gc_filt)
dev <- computeDeviations(object = peak_count, annotations = motif_ix
                         #,background_peaks = bg
                         )

dev_scores <- deviations(dev) %>%
  as.data.frame() %>%
  rownames_to_column(var = "motif_id") %>%
  left_join(motifs_selected) %>%
  pivot_longer(cols = -c(motif_id, tf), names_to = "sample", values_to = "chromVAR_zscore") %>%
  mutate(replicate = gsub(".*(.[1-3]{1}$)", "\\1", sample)) %>%
  mutate(condition = gsub("_[1-3]{1}$", "", sample)) %>%
  mutate(chromVAR_zscore_mean = ave(chromVAR_zscore, motif_id, condition, FUN = mean)) %>%
  distinct(motif_id, tf, condition, replicate, chromVAR_zscore_mean, chromVAR_zscore) %>%
  filter(condition %in% c("SOX2_C11_24H", "SOX2_C11_0H")) %>%
  dplyr::select(-replicate, -chromVAR_zscore) %>%
  distinct()

dev_scores_ctrl <- dev_scores %>%
  filter(condition %in% c("SOX2_C11_0H")) %>%
  dplyr::select("ref_condition" = condition, "ctrl_zscore" = chromVAR_zscore_mean, tf) 

condition_match <- data.frame(condition = c("SOX2_C11_24H"),
                              ref_condition = c("SOX2_C11_0H"))

dev_scores_dif <- dev_scores %>%
  left_join(condition_match) %>%
  left_join(dev_scores_ctrl) %>%
  mutate(chromVAR_zscore_dif_mean = chromVAR_zscore_mean - ctrl_zscore) %>%
  na.omit()


saveRDS(dev_scores_dif, "/DATA/shared/projects/TF_REPORTERS/ATAC_TF_reporter_comparison/data/chromVAR/dev_scores_teun.rds")
```






# Session Info
```{r}
paste("Run time: ",format(Sys.time()-StartTime))
getwd()
date()
sessionInfo()
```

