---
title: "Multiplexed identification of TF-specific reporters"
author: 
  - name: "Max Trauernicht"
    email: "m.trauernicht@nki.nl"
    affiliation: "Netherlands Cancer Institute - van Steensel lab"

date: '`r format(Sys.time(), "%d/%m/%Y")`'
output:
  html_document:
    theme: united
    highlight: pygments
    fig_caption: yes
    code_folding: hide
    df_print: kable
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
---

---

### Aim
~36,000 reporters for 86 TFs were transfected into 9 different cell types + treated with ~90 different siRNAs or various signaling molecules. In this script I will analyze TF reporter activities in detail and review how individual reporters respond to TF concentration variations.

---

## Setup {.tabset}

<!-- little HTML script to do indentation of the table of contents -->
<script>
    $(document).ready(function() {
      $items = $('div#TOC li');
      $items.each(function(idx) {
        num_ul = $(this).parentsUntil('#TOC').length;
        $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
      });

    });
</script>

```{css, echo = FALSE}
div.sourceCode {
  overflow-x: hidden;
}
```

### Libraries 

```{r setup, out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
knitr::opts_chunk$set(echo = TRUE)
StartTime <-Sys.time()

# 8-digit Date tag:
Date <- substr(gsub("-","",Sys.time()),1,8) 
# libraries:
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(tibble)
library(pheatmap)
library(visNetwork)
library(ggbeeswarm)
library(ggforce)
library(viridis)
library(plyr)
library(igraph)
library(ggraph)
library(cowplot)
library(gridExtra)
library(pROC)
library(tidyr)
library(stringr)
library(ggrastr)
library(readr)
library(ggbiplot)
library(ggh4x)
library(ggiraph)
library(plotly)
library(AMR)
library(umap)
library(ggrepel)
library(chromVAR)
library(Matrix)
library(SummarizedExperiment)
library(BiocParallel)
library(BSgenome.Mmusculus.UCSC.mm10)
library(BSgenome.Hsapiens.UCSC.hg38)
library(universalmotif)
library(motifmatchr)
library(limma)
```

### Functions

```{r out.width= "80%", fig.align= "center"}
### Custom functions
SetFileName <- function(filename, initials) {
  # Set filename with extension and initials to make filename with date integrated.
  filename <- substitute(filename)
  initials <- substitute(initials)
  filename <- paste0(initials, Date, filename)
  filename
}


# Extract p-value from linear model
lmp <- function (modelobject) {
    if (class(modelobject) != "lm") stop("Not an object of class 'lm' ")
    f <- summary(modelobject)$fstatistic
    p <- pf(f[1],f[2],f[3],lower.tail=F)
    attributes(p) <- NULL
    return(p)
}
```

### Loading ATAC-seq data - mouse cells

```{r out.width= "80%", fig.align= "center"}
# Import peak count information for ChromVAR
peak_count <- readRDS("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/ATAC_seq_analyses/rds/experiment_7784_20250206.rds")
assayNames(peak_count) <- "counts"

# Filter samples
#peak_count <- peak_count[, !peak_count@colData$run %in% c(8010, 8011)] ## Run separately for each seq run
peak_count <- peak_count[, peak_count@colData$celltype == "mES"] 
peak_count <- peak_count[, peak_count@colData$treatment %in% c("2i_LIF", "NT", "CH", "POU5F1_DEG", "POU5F1_ctrl", "SOX2_DEG",
                                                               "SOX2_ctrl", "FBS", "FK", "Heat", "LIF", "LIF_1h", "LIF_3h",
                                                               "LIF_6h", "TFCP2L1", "FOXA1_OE", "FOXA1_ctrl")] 
peak_count <- peak_count[, peak_count@colData$basename != "1_mES_r3_TAAGGCGA-AGGCTTAG_S1"] 

min_samples <- 1  # Adjust as needed
valid_groups <- names(which(table(peak_count@colData$treatment) >= min_samples))
peak_count <- peak_count[, peak_count@colData$treatment %in% valid_groups]


# Add GC bias
peak_count <- addGCBias(peak_count, genome = BSgenome.Mmusculus.UCSC.mm10)

## Filter peaks - each peak must have at least one count in one condition
peak_count <- filterPeaks(sort(peak_count), non_overlapping = TRUE)

## Extract metadata
metadata <- data.frame(peak_count@colData) %>%
  rownames_to_column("sample") %>%
  distinct(sample, run, celltype, treatment, replicate)


# Import motif information
seq_logos <- readRDS("/DATA/usr/m.trauernicht/projects/SuRE-TF/data/library_design/lambert_2018/seq_logos.rds")
motifs_selected <- readRDS("/DATA/usr/m.trauernicht/projects/SuRE-TF/data/library_design/lambert_2018/motifs_selected.rds") %>%
  mutate(tf = toupper(tf)) %>%
  mutate(tf = ifelse(tf == "AHR:ARNT", "AHR::ARNT", tf)) %>%
  mutate(tf = ifelse(tf == "RAR:RXR", "RARA:RXRA", tf)) %>%
  rbind(data.frame(tf = c("STAT1", "STAT4", "ATF2", "ATF4", "ATF6", "CLOCK"),
                   motif_id = c("M06794_1.94d.txt", "M06797_1.94d.txt", "M08481_1.94d.txt", "M08489_1.94d.txt", "M09435_1.94d.txt", "M05596_1.94.txt")))
  
seq_logos_pwm <- list()
for (i in names(seq_logos)) {
  if (i %in% motifs_selected$motif_id) {
  seq_logos_pwm[[i]] <- universalmotif::convert_motifs(seq_logos[[i]], class = "TFBSTools-PWMatrix")
  seq_logos_pwm[[i]]@name <- i
  seq_logos_pwm[[i]]@strand <- "+-"
  }
}
# 
# seq_logos_pwm <- list()
# for (i in names(seq_logos)) {
#   seq_logos_pwm[[i]] <- universalmotif::convert_motifs(seq_logos[[i]], class = "TFBSTools-PWMatrix")
#   name_tf <- all_motifs$tf[all_motifs$motif_id == i]
#   seq_logos_pwm[[i]]@name <- name_tf
#   seq_logos_pwm[[i]]@strand <- "+-"
# }
# 

seq_logos_pwm_all <- do.call(TFBSTools::PWMatrixList, seq_logos_pwm)

motif_ix <- matchMotifs(seq_logos_pwm_all, peak_count, 
                        genome = BSgenome.Mmusculus.UCSC.mm10)
```

---

### Run chromVAR
```{r out.width= "80%", fig.align= "center"}
# Compute deviations
bg <- getBackgroundPeaks(object = peak_count, niterations = 2000)
dev <- computeDeviations(object = peak_count, annotations = motif_ix
                         ,background_peaks = bg
                         )

dev_scores <- deviations(dev) %>%
  as.data.frame() %>%
  rownames_to_column(var = "motif_id") %>%
  left_join(motifs_selected) %>%
  pivot_longer(cols = -c(motif_id, tf), names_to = "sample", values_to = "chromVAR_zscore") %>%
  # mutate(replicate = gsub(".*(R[1-3]{1}$)", "\\1", sample)) %>%
  mutate(condition = gsub("_R[1-4]{1}$", "", sample)) %>%
  left_join(metadata, by = "sample") %>%
  mutate(sample = gsub("\\.1$", "", sample)) %>%
  mutate(sample = paste(sample, run, sep = "_")) %>%
  mutate(chromVAR_zscore_mean = ave(chromVAR_zscore, motif_id, condition, FUN = mean)) %>%
  distinct(motif_id, tf, condition, run, replicate, chromVAR_zscore_mean, chromVAR_zscore)

# dev_scores_ctrl <- dev_scores %>%
#   filter(condition %in% c("mES_2i_LIF", "mES_FOXA1_ctrl", "mES_POU5F1_ctrl", "mES_SOX2_ctrl", "mES_NT")) %>%
#   mutate(replicate = as.character(replicate)) %>%
#   dplyr::select("ref_condition" = condition, "ctrl_zscore" = chromVAR_zscore, tf, replicate, run) %>%
#   mutate(replicate = ifelse(ref_condition == "NPC_DMSO", "R1", replicate)) %>%
#   mutate(run = ifelse(ref_condition == "NPC_DMSO", 8010, run)) %>%
#   mutate(replicate = as.factor(replicate)) 


condition_match <- data.frame(condition = c("mES_LIF", "mES_CH", "mES_FOXA1_OE", "mES_POU5F1_DEG", "mES_SOX2_DEG", "mES_TFCP2L1", 
                                            "mES_LIF_1h", "mES_LIF_3h", "mES_LIF_6h", "mES_FBS", 
                                            "mES_FK", "mES_Heat"),
                              ref_condition = c("mES_2i_LIF",  "mES_2i_LIF", "mES_FOXA1_ctrl", "mES_POU5F1_ctrl", "mES_SOX2_ctrl", "mES_NT",
                                                "mES_NT", "mES_NT", "mES_NT", "mES_NT",
                                                "mES_NT", "mES_NT"))

# dev_scores_dif <- dev_scores %>%
#   left_join(condition_match) %>%
#   left_join(dev_scores_ctrl) %>%
#   mutate(chromVAR_zscore_dif = chromVAR_zscore - ctrl_zscore) %>%
#   mutate(chromVAR_zscore_dif_mean = ave(chromVAR_zscore_dif, tf, condition, FUN = function (x) mean(x, na.rm = T))) %>%
#   filter(condition %in% c("mES_CH", "mES_LIF", "mES_POU5F1_DEG", "mES_SOX2_DEG", "mES_FBS", "mES_FK", "mES_Heat",
#                           "mES_LIF_1h", "mES_LIF_3h", "mES_LIF_6h", "mES_TFCP2L1", "mES_FOXA1_OE"))

# # Calculate differential deviation p-values
# dev_dif_ch <- differentialDeviations(dev[,dev@colData$treatment %in% c("2i_LIF", "CH")], group = "treatment") %>%
#   rownames_to_column(var = "motif_id") %>%
#   distinct(motif_id, p_value) %>%
#   mutate(condition = "mES_CH")
# 
# dev_dif_lif <- differentialDeviations(dev[,dev@colData$treatment %in% c("2i_LIF", "LIF")], group = "treatment") %>%
#   rownames_to_column(var = "motif_id") %>%
#   distinct(motif_id, p_value) %>%
#   mutate(condition = "mES_LIF")
# 
# dev_dif_heat <- differentialDeviations(dev[,dev@colData$treatment %in% c("NT", "Heat")], group = "treatment") %>%
#   rownames_to_column(var = "motif_id") %>%
#   distinct(motif_id, p_value) %>%
#   mutate(condition = "mES_Heat")%>%
#   left_join(motifs_selected) 
# 
# dev_dif_pou5f1 <- differentialDeviations(dev[,dev@colData$treatment %in% c("POU5F1_ctrl", "POU5F1_DEG")], group = "treatment") %>%
#   rownames_to_column(var = "motif_id") %>%
#   distinct(motif_id, p_value) %>%
#   mutate(condition = "mES_POU5F1_DEG")
# 
# dev_dif_sox2 <- differentialDeviations(dev[,dev@colData$treatment %in% c("SOX2_ctrl", "SOX2_DEG")], group = "treatment") %>%
#   rownames_to_column(var = "motif_id") %>%
#   distinct(motif_id, p_value) %>%
#   mutate(condition = "mES_SOX2_DEG")
# 
# dev_dif_fbs <- differentialDeviations(dev[,dev@colData$treatment %in% c("NT", "FBS")], group = "treatment") %>%
#   rownames_to_column(var = "motif_id") %>%
#   distinct(motif_id, p_value) %>%
#   mutate(condition = "mES_FBS")
# 
# dev_dif_fk <- differentialDeviations(dev[,dev@colData$treatment %in% c("NT", "FK")], group = "treatment") %>%
#   rownames_to_column(var = "motif_id") %>%
#   distinct(motif_id, p_value) %>%
#   mutate(condition = "mES_FK")
# 
# dev_dif_lif_1h <- differentialDeviations(dev[,dev@colData$treatment %in% c("NT", "LIF_1h")], group = "treatment") %>%
#   rownames_to_column(var = "motif_id") %>%
#   distinct(motif_id, p_value) %>%
#   mutate(condition = "mES_LIF_1h")
# 
# dev_dif_lif_3h <- differentialDeviations(dev[,dev@colData$treatment %in% c("NT", "LIF_3h")], group = "treatment") %>%
#   rownames_to_column(var = "motif_id") %>%
#   distinct(motif_id, p_value) %>%
#   mutate(condition = "mES_LIF_3h")
# 
# dev_dif_lif_6h <- differentialDeviations(dev[,dev@colData$treatment %in% c("NT", "LIF_6h")], group = "treatment") %>%
#   rownames_to_column(var = "motif_id") %>%
#   distinct(motif_id, p_value) %>%
#   mutate(condition = "mES_LIF_6h")
# 
# dev_dif_tfcp2l1 <- differentialDeviations(dev[,dev@colData$treatment %in% c("NT", "TFCP2L1")], group = "treatment") %>%
#   rownames_to_column(var = "motif_id") %>%
#   distinct(motif_id, p_value) %>%
#   mutate(condition = "mES_TFCP2L1")
# 
# # dev_dif_foxa1 <- differentialDeviations(dev[,dev@colData$treatment %in% c("FOXA1_ctrl", "FOXA1_OE")], group = "treatment") %>%
# #   rownames_to_column(var = "motif_id") %>%
# #   distinct(motif_id, p_value_adjusted) %>%
# #   mutate(condition = "mES_FOXA1_OE")
# 
# dev_dif <- rbind(dev_dif_ch, dev_dif_lif, dev_dif_heat, dev_dif_pou5f1, dev_dif_sox2, dev_dif_fbs, dev_dif_fk,
#                  dev_dif_lif_1h, dev_dif_lif_3h, dev_dif_lif_6h, dev_dif_tfcp2l1)



devMat <- assays(dev)$deviations

# Estimate the fold changes and standard errors by fitting a linear model for each gene

condition <- factor(colData(dev)$treatment)
design <- model.matrix(~ 0 + condition)
colnames(design) <- make.names(levels(condition))

# Fit model once to all data
fit <- lmFit(devMat, design)

# List of named contrasts — use clean condition names (no quotes)
contrast_list <- list(
  `CH_vs_2i_LIF` = c("CH", "2i_LIF"),
  `LIF_vs_2i_LIF` = c("LIF", "NT"),
  `Heat_vs_NT` = c("Heat", "NT"),
  `POU5F1_DEG_vs_ctrl` = c("POU5F1_DEG", "POU5F1_ctrl"),
  `SOX2_DEG_vs_ctrl` = c("SOX2_DEG", "SOX2_ctrl"),
  `FBS_vs_NT` = c("FBS", "NT"),
  `FK_vs_NT` = c("FK", "NT"),
  `LIF_1h_vs_NT` = c("LIF_1h", "NT"),
  `LIF_3h_vs_NT` = c("LIF_3h", "NT"),
  `LIF_6h_vs_NT` = c("LIF_6h", "NT"),
  `TFCP2L1_vs_NT` = c("TFCP2L1", "NT"),
  `FOXA1_OE_vs_ctrl` = c("FOXA1_OE", "FOXA1_ctrl")
)

# Run contrasts in a loop
results_list <- list()

results_list[[i]] <- for (i in 1:12) {
  
  print(i)
  
  groups <- contrast_list[[i]]
  contrast_names <- make.names(groups)

  # Skip if contrast levels don't exist in design
  if (!all(contrast_names %in% colnames(design))) {
    warning(paste("Skipping contrast", name, "— missing levels"))
    return(NULL)
  }

  # Create contrast matrix for this comparison
  contrast_matrix <- makeContrasts(
    contr = paste0(contrast_names[1], " - ", contrast_names[2]),
    levels = design
  )

  # Fit and apply eBayes for this contrast
  fit_contrast <- contrasts.fit(fit, contrast_matrix)
  fit_contrast <- eBayes(fit_contrast)

  # Extract table of results
  top <- topTable(fit_contrast, number = Inf)
  top$motif_id <- rownames(top)
  top$group1 <- groups[1]
  top$group2 <- groups[2]

  results_list[[i]] <- top
}

# Combine into one dataframe
combined_top_table <- bind_rows(results_list) %>%
  left_join(motifs_selected) %>%
  dplyr::select(tf, "chromVAR_zscore_dif_mean" = logFC, "p_value_adjusted" = adj.P.Val, "condition" = group1) 


dev_scores_dif <- dev_scores_dif %>%
  left_join(dev_dif)

saveRDS(dev_scores_dif, "/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/data/chromVAR/dev_scores_mouse_2025_mouse_all.rds")
```

---



## Run on individual comparisons instead of all together

```{r out.width= "80%", fig.align= "center"}
# Import peak count information for ChromVAR
peak_count <- readRDS("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/ATAC_seq_analyses/rds/experiment_7784_20250206.rds")
assayNames(peak_count) <- "counts"

# Filter samples
peak_count <- peak_count[, peak_count@colData$celltype == "mES"] ## Run seperately for each cell type
peak_count <- peak_count[, peak_count@colData$treatment %in% c("2i_LIF", "FK")] ## Run seperately for each cell type
peak_count <- peak_count[, !peak_count@colData$basename == "1_mES_r3_TAAGGCGA-AGGCTTAG_S1"] ## Run seperately for each cell type


# Add GC bias
peak_count <- addGCBias(peak_count, genome = BSgenome.Mmusculus.UCSC.mm10)

## Filter peaks - each peak must have at least one count in one condition
peak_count <- filterPeaks(sort(peak_count), non_overlapping = TRUE)

## Extract metadata
metadata <- data.frame(peak_count@colData) %>%
  rownames_to_column("sample") %>%
  distinct(sample, run, celltype, treatment, replicate)


# Import motif information
seq_logos <- readRDS("/DATA/usr/m.trauernicht/projects/SuRE-TF/data/library_design/lambert_2018/seq_logos.rds")
motifs_selected <- readRDS("/DATA/usr/m.trauernicht/projects/SuRE-TF/data/library_design/lambert_2018/motifs_selected.rds") %>%
  mutate(tf = toupper(tf)) %>%
  mutate(tf = ifelse(tf == "AHR:ARNT", "AHR::ARNT", tf)) %>%
  mutate(tf = ifelse(tf == "RAR:RXR", "RARA:RXRA", tf)) %>%
  rbind(data.frame(tf = c("STAT1", "STAT4", "ATF2", "ATF4", "ATF6", "CLOCK"),
                   motif_id = c("M06794_1.94d.txt", "M06797_1.94d.txt", "M08481_1.94d.txt", "M08489_1.94d.txt", "M09435_1.94d.txt", "M05596_1.94.txt")))
  
seq_logos_pwm <- list()
for (i in names(seq_logos)) {
  if (i %in% motifs_selected$motif_id) {
  seq_logos_pwm[[i]] <- universalmotif::convert_motifs(seq_logos[[i]], class = "TFBSTools-PWMatrix")
  seq_logos_pwm[[i]]@name <- i
  seq_logos_pwm[[i]]@strand <- "+-"
  }
}
# 
# seq_logos_pwm <- list()
# for (i in names(seq_logos)) {
#   seq_logos_pwm[[i]] <- universalmotif::convert_motifs(seq_logos[[i]], class = "TFBSTools-PWMatrix")
#   name_tf <- all_motifs$tf[all_motifs$motif_id == i]
#   seq_logos_pwm[[i]]@name <- name_tf
#   seq_logos_pwm[[i]]@strand <- "+-"
# }
# 

seq_logos_pwm_all <- do.call(TFBSTools::PWMatrixList, seq_logos_pwm)

motif_ix <- matchMotifs(seq_logos_pwm_all, peak_count, 
                        genome = BSgenome.Mmusculus.UCSC.mm10)
```

---

### Run chromVAR
```{r out.width= "80%", fig.align= "center"}
# Compute deviations
bg <- getBackgroundPeaks(object = peak_count, niterations = 2000)
dev <- computeDeviations(object = peak_count, annotations = motif_ix
                         ,background_peaks = bg
                         )

dev_dif <- differentialDeviations(dev, group = "treatment") %>%
  rownames_to_column(var = "motif_id") %>%
  left_join(motifs_selected) 

dev_scores <- deviations(dev) %>%
  as.data.frame() %>%
  rownames_to_column(var = "motif_id") %>%
  left_join(motifs_selected) %>%
  pivot_longer(cols = -c(motif_id, tf), names_to = "sample", values_to = "chromVAR_zscore") %>%
  # mutate(replicate = gsub(".*(R[1-3]{1}$)", "\\1", sample)) %>%
  mutate(condition = gsub("_R[1-3]{1}$", "", sample)) %>%
  left_join(metadata, by = "sample") %>%
  mutate(sample = gsub("\\.1$", "", sample)) %>%
  mutate(sample = paste(sample, run, sep = "_")) %>%
  mutate(chromVAR_zscore_mean = ave(chromVAR_zscore, motif_id, condition, FUN = mean)) %>%
  distinct(motif_id, tf, condition, run, replicate, chromVAR_zscore_mean, chromVAR_zscore)

dev_scores_ctrl <- dev_scores %>%
  filter(condition %in% c("mES_2i_LIF", "mES_FOXA1_ctrl", "mES_POU5F1_ctrl", "mES_SOX2_ctrl", "mES_NT", "NPC_DMSO")) %>%
  mutate(replicate = as.character(replicate)) %>%
  dplyr::select("ref_condition" = condition, "ctrl_zscore" = chromVAR_zscore, tf, replicate, run) %>%
  mutate(replicate = ifelse(ref_condition == "NPC_DMSO", "R1", replicate)) %>%
  mutate(run = ifelse(ref_condition == "NPC_DMSO", 8010, run)) %>%
  mutate(replicate = as.factor(replicate)) 

dev_scores_ctrl2 <- dev_scores_ctrl %>%
  filter(ref_condition == "NPC_DMSO") %>%
  mutate(run = 7822)

dev_scores_ctrl <- rbind(dev_scores_ctrl, dev_scores_ctrl2)


condition_match <- data.frame(condition = c("mES_LIF", "mES_PD", "mES_CH", "mES_FOXA1_OE", "mES_POU5F1_DEG", "mES_SOX2_DEG", "mES_TFCP2L1", "mES_SP1",
                                            "mES_LIF_1h", "mES_LIF_3h", "mES_LIF_6h", "mES_HQ_1h", "mES_HQ_3h", "mES_HQ_6h", "mES_10PD", "mES_FBS", 
                                            "mES_FK", "mES_vitC", "mES_Heat", "mES_PMA", "mES_Nutlin", "mES_FK_new", "NPC_HQ", "NPC_cGAMP", "NPC_FK",
                                            "mES_2i_LIF"),
                              ref_condition = c("mES_2i_LIF", "mES_2i_LIF", "mES_2i_LIF", "mES_FOXA1_ctrl", "mES_POU5F1_ctrl", "mES_SOX2_ctrl", "mES_NT",
                                                "mES_NT",
                                                "mES_NT", "mES_NT", "mES_NT", "mES_NT", "mES_NT", "mES_NT", "mES_NT", "mES_NT",
                                                "mES_NT", "mES_NT", "mES_NT","mES_NT", "mES_NT", "mES_NT", "NPC_DMSO", "NPC_DMSO", "NPC_DMSO", "NPC_DMSO"))

dev_scores_dif <- dev_scores %>%
  left_join(condition_match) %>%
  left_join(dev_scores_ctrl) %>%
  mutate(chromVAR_zscore_dif = chromVAR_zscore - ctrl_zscore) %>%
  mutate(chromVAR_zscore_dif_mean = ave(chromVAR_zscore_dif, tf, condition, FUN = function (x) mean(x, na.rm = T)))

# Compute variability
variability <- computeVariability(dev) %>%
  rownames_to_column(var = "motif_id") %>%
  left_join(motifs_selected) 

plotVariability(variability, use_plotly = FALSE)

#saveRDS(dev_scores_dif, "/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/data/chromVAR/dev_scores_mouse_2025_mouse_all.rds")


devMat <- assays(dev)$z

# Estimate the fold changes and standard errors by fitting a linear model for each gene

condition <- factor(colData(dev)$treatment)
design <- model.matrix(~ 0 + condition)
colnames(design) <- levels(condition)

fit <- lmFit(devMat, 
             design)

# Apply empirical Bayes smoothing to the standard errors

fit <- eBayes(fit)

# Show statistics for the top 10 genes

topTFs <- topTable(fit, number = Inf) %>%
  rownames_to_column(var = "motif_id") %>%
  left_join(motifs_selected) 
  
```











## ChromVAR on the human conditions

### Loading data

```{r out.width= "80%", fig.align= "center"}
# Import peak count information for ChromVAR
peak_count <- readRDS("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/ATAC_seq_analyses/rds/experiment_7784_20241003_human.rds")
assayNames(peak_count) <- "counts"

# Add GC bias
peak_count <- addGCBias(peak_count, genome = BSgenome.Hsapiens.UCSC.hg38)

## Filter peaks - each peak must have at least one count in one condition
peak_count <- filterPeaks(GenomicRanges::sort(peak_count), non_overlapping = TRUE)

motif_ix <- matchMotifs(seq_logos_pwm_all, peak_count, 
                        genome = BSgenome.Hsapiens.UCSC.hg38)
```

---

### Run chromVAR
```{r out.width= "80%", fig.align= "center"}
# Compute deviations
#bg <- getBackgroundPeaks(object = sox2_peak_count_gc_filt)
dev <- computeDeviations(object = peak_count, annotations = motif_ix
                         #,background_peaks = bg
                         )

dev_dif <- differentialDeviations(dev, group = peak_count@colData$treatment)

dev_scores <- deviationScores(dev) %>%
  as.data.frame() %>%
  rownames_to_column(var = "motif_id") %>%
  left_join(motifs_selected) %>%
  pivot_longer(cols = -c(motif_id, tf), names_to = "sample", values_to = "chromVAR_zscore") %>%
  mutate(replicate = gsub(".*(.[1-3]{1}$)", "\\1", sample)) %>%
  mutate(condition = gsub("_.[1-3]{1}$", "", sample)) %>%
  mutate(chromVAR_zscore_mean = ave(chromVAR_zscore, motif_id, condition, FUN = mean)) %>%
  distinct(motif_id, tf, condition, replicate, chromVAR_zscore_mean, chromVAR_zscore)

dev_scores_ctrl <- dev_scores %>%
  filter(condition %in% c("U2OS_DMSO", "MCF7_DMSO")) %>%
  dplyr::select("ref_condition" = condition, "ctrl_zscore" = chromVAR_zscore, tf, replicate) 

condition_match <- data.frame(condition = c("U2OS_PMA", "U2OS_Heat", "U2OS_Calcitriol", "MCF7_Hexestrol"),
                              ref_condition = c("U2OS_DMSO", "U2OS_DMSO", "U2OS_DMSO", "MCF7_DMSO"))

dev_scores_dif <- dev_scores %>%
  left_join(condition_match) %>%
  left_join(dev_scores_ctrl) %>%
  mutate(chromVAR_zscore_dif = chromVAR_zscore - ctrl_zscore) %>%
  mutate(chromVAR_zscore_dif_mean = ave(chromVAR_zscore_dif, tf, condition, FUN = mean))

# Compute variability
variability <- computeVariability(dev)

plotVariability(variability, use_plotly = FALSE)

saveRDS(dev_scores_dif, "/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/data/chromVAR/dev_scores_human.rds")
```






# Session Info
```{r}
paste("Run time: ",format(Sys.time()-StartTime))
getwd()
date()
sessionInfo()
```

