---
title: "Multiplexed identification of TF-specific reporters"
author: 
  - name: "Max Trauernicht"
    email: "m.trauernicht@nki.nl"
    affiliation: "Netherlands Cancer Institute - van Steensel lab"

date: '`r format(Sys.time(), "%d/%m/%Y")`'
output:
  html_document:
    theme: united
    highlight: pygments
    fig_caption: yes
    code_folding: hide
    df_print: kable
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
---

---

### Aim
~36,000 reporters for 86 TFs were transfected into 9 different cell types + treated with ~90 different siRNAs or various signaling molecules. In this script I will analyze TF reporter activities in detail and review how individual reporters respond to TF concentration variations.

---

## Setup {.tabset}

<!-- little HTML script to do indentation of the table of contents -->
<script>
    $(document).ready(function() {
      $items = $('div#TOC li');
      $items.each(function(idx) {
        num_ul = $(this).parentsUntil('#TOC').length;
        $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
      });

    });
</script>

```{css, echo = FALSE}
div.sourceCode {
  overflow-x: hidden;
}
```

### Libraries 

```{r setup, out.width= "80%", fig.align= "center", warning = FALSE, message= FALSE}
knitr::opts_chunk$set(echo = TRUE)
StartTime <-Sys.time()

# 8-digit Date tag:
Date <- substr(gsub("-","",Sys.time()),1,8) 
# libraries:
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(maditr)
library(tibble)
library(pheatmap)
library(ggpubr)
library(visNetwork)
library(ggbeeswarm)
library(ggforce)
library(viridis)
library(plyr)
library(igraph)
library(ggraph)
library(cowplot)
library(gridExtra)
library(pROC)
library(tidyr)
library(stringr)
library(randomForest)
library(ggrastr)
library(readr)
library(ggbiplot)
library(IHW)
library(biomaRt)
library(ggh4x)
library(ggiraph)
library(plotly)
library(AMR)
library(umap)
library(ggrepel)
library(chromVAR)
library(motifmatchr)
library(Matrix)
library(SummarizedExperiment)
library(BiocParallel)
library(JASPAR2016)
library(BSgenome.Mmusculus.UCSC.mm10)
library(BSgenome.Hsapiens.UCSC.hg38)
```

### Functions

```{r out.width= "80%", fig.align= "center"}
### Custom functions
SetFileName <- function(filename, initials) {
  # Set filename with extension and initials to make filename with date integrated.
  filename <- substitute(filename)
  initials <- substitute(initials)
  filename <- paste0(initials, Date, filename)
  filename
}


# Extract p-value from linear model
lmp <- function (modelobject) {
    if (class(modelobject) != "lm") stop("Not an object of class 'lm' ")
    f <- summary(modelobject)$fstatistic
    p <- pf(f[1],f[2],f[3],lower.tail=F)
    attributes(p) <- NULL
    return(p)
}

# Set custom ggplot2 theme and custom colors
theme_classic_lines <- function() {
  theme_pubr(border = F, legend = "top") +
            theme(panel.grid.major = element_line(colour = "#adb5bd", linewidth = 0.25),
                  strip.background = element_rect(fill = "#ced4da")
            )
}

theme_classic_lines_45 <- function() {
  theme_pubr(border = T, legend = "top", x.text.angle = 45) +
            theme(panel.grid.major = element_line(colour = "#adb5bd", linewidth = 0.25),
                  strip.background = element_rect(fill = "#ced4da")
            )
}

theme_classic_lines_90 <- function() {
  theme_pubr(border = T, legend = "top", x.text.angle = 90) +
            theme(panel.grid.major = element_line(colour = "#adb5bd", linewidth = 0.25),
                  strip.background = element_rect(fill = "#ced4da")
            )
}

theme_barplot <- function() {
  theme_pubr(border = T, legend = "none") +
            theme(panel.grid.major.x = element_line(colour = "black", linewidth = 0.25),
                  strip.background = element_blank(),
                  strip.text = element_text(face="bold", hjust=0)
            )
}

theme_set(theme_classic_lines())

colors_diverse <- c("#264653", "#9AC1AE", "#5D987B", "#f2cc8f", "#e76f51")

cell_colors <- c("A549" = "#f2cc8f", "HCT116" = "#ED1C24", "HEK293" = "#00B4D8", 
                               "HepG2" = "#B3B3B3", "K562" = "#A67C52", "MCF7" = "#81B29A", 
                               "U2OS" = "#3D405B", "mES" = "#EAB69F", "NPC" = "#E07A5F")

ggplot_custom <- function(...) ggplot2::ggplot(...) + 
  scale_color_manual(values = colors_diverse) + 
  scale_fill_manual(values = colors_diverse)
```

### Loading ATAC-seq data - mouse cells

```{r out.width= "80%", fig.align= "center"}
# Import peak count information for ChromVAR
peak_count <- readRDS("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/ATAC_seq_analyses/rds/experiment_7784_20250206.rds")
assayNames(peak_count) <- "counts"

# Filter samples
#peak_count <- peak_count[, !peak_count@colData$run %in% c(8010, 8011)] ## Run separately for each seq run
#peak_count <- peak_count[, peak_count@colData$celltype == "mES"] ## Run seperately for each cell type
min_samples <- 2  # Adjust as needed
valid_groups <- names(which(table(peak_count@colData$treatment) >= min_samples))
peak_count <- peak_count[, peak_count@colData$treatment %in% valid_groups]

# Add GC bias
peak_count <- addGCBias(peak_count, genome = BSgenome.Mmusculus.UCSC.mm10)

## Filter peaks - each peak must have at least one count in one condition
peak_count <- filterPeaks(sort(peak_count), non_overlapping = TRUE)

## Extract metadata
metadata <- data.frame(peak_count@colData) %>%
  rownames_to_column("sample") %>%
  distinct(sample, run, celltype, treatment, replicate)


# Import motif information
seq_logos <- readRDS("/DATA/usr/m.trauernicht/projects/SuRE-TF/data/library_design/lambert_2018/seq_logos.rds")
motifs_selected <- readRDS("/DATA/usr/m.trauernicht/projects/SuRE-TF/data/library_design/lambert_2018/motifs_selected.rds") %>%
  mutate(tf = toupper(tf)) %>%
  mutate(tf = ifelse(tf == "AHR:ARNT", "AHR::ARNT", tf)) %>%
  mutate(tf = ifelse(tf == "RAR:RXR", "RARA:RXRA", tf)) %>%
  rbind(data.frame(tf = c("STAT1", "STAT4", "ATF2", "ATF4", "ATF6", "CLOCK"),
                   motif_id = c("M06794_1.94d.txt", "M06797_1.94d.txt", "M08481_1.94d.txt", "M08489_1.94d.txt", "M09435_1.94d.txt", "M05596_1.94.txt")))
  
seq_logos_pwm <- list()
for (i in names(seq_logos)) {
  if (i %in% motifs_selected$motif_id) {
  seq_logos_pwm[[i]] <- universalmotif::convert_motifs(seq_logos[[i]], class = "TFBSTools-PWMatrix")
  seq_logos_pwm[[i]]@name <- i
  seq_logos_pwm[[i]]@strand <- "+-"
  }
}
# 
# seq_logos_pwm <- list()
# for (i in names(seq_logos)) {
#   seq_logos_pwm[[i]] <- universalmotif::convert_motifs(seq_logos[[i]], class = "TFBSTools-PWMatrix")
#   name_tf <- all_motifs$tf[all_motifs$motif_id == i]
#   seq_logos_pwm[[i]]@name <- name_tf
#   seq_logos_pwm[[i]]@strand <- "+-"
# }
# 

seq_logos_pwm_all <- do.call(TFBSTools::PWMatrixList, seq_logos_pwm)

motif_ix <- matchMotifs(seq_logos_pwm_all, peak_count, 
                        genome = BSgenome.Mmusculus.UCSC.mm10)
```

---

### Run chromVAR
```{r out.width= "80%", fig.align= "center"}
# Compute deviations
#bg <- getBackgroundPeaks(object = sox2_peak_count_gc_filt)
dev <- computeDeviations(object = peak_count, annotations = motif_ix
                         #,background_peaks = bg
                         )

dev_dif <- differentialDeviations(dev, group = peak_count@colData$treatment)

dev_scores <- deviationScores(dev) %>%
  as.data.frame() %>%
  rownames_to_column(var = "motif_id") %>%
  left_join(motifs_selected) %>%
  pivot_longer(cols = -c(motif_id, tf), names_to = "sample", values_to = "chromVAR_zscore") %>%
  # mutate(replicate = gsub(".*(R[1-3]{1}$)", "\\1", sample)) %>%
  mutate(condition = gsub("_R[1-3]{1}$", "", sample)) %>%
  left_join(metadata, by = "sample") %>%
  mutate(sample = gsub("\\.1$", "", sample)) %>%
  mutate(sample = paste(sample, run, sep = "_")) %>%
  mutate(chromVAR_zscore_mean = ave(chromVAR_zscore, motif_id, condition, FUN = mean)) %>%
  distinct(motif_id, tf, condition, run, replicate, chromVAR_zscore_mean, chromVAR_zscore)

dev_scores_ctrl <- dev_scores %>%
  filter(condition %in% c("mES_2i_LIF", "mES_FOXA1_ctrl", "mES_POU5F1_ctrl", "mES_SOX2_ctrl", "mES_NT", "NPC_DMSO")) %>%
  mutate(replicate = as.character(replicate)) %>%
  dplyr::select("ref_condition" = condition, "ctrl_zscore" = chromVAR_zscore, tf, replicate, run) %>%
  mutate(replicate = ifelse(ref_condition == "NPC_DMSO", "R1", replicate)) %>%
  mutate(run = ifelse(ref_condition == "NPC_DMSO", 8010, run)) %>%
  mutate(replicate = as.factor(replicate)) 

condition_match <- data.frame(condition = c("mES_LIF", "mES_PD", "mES_CH", "mES_FOXA1_OE", "mES_POU5F1_DEG", "mES_SOX2_DEG", "mES_TFCP2L1", "mES_SP1",
                                            "mES_LIF_1h", "mES_LIF_3h", "mES_LIF_6h", "mES_HQ_1h", "mES_HQ_3h", "mES_HQ_6h", "mES_10PD", "mES_FBS", 
                                            "mES_FK", "mES_vitC", "mES_Heat", "mES_PMA", "mES_Nutlin", "mES_FK_new", "NPC_HQ", "NPC_cGAMP", "NPC_FK"),
                              ref_condition = c("mES_2i_LIF", "mES_2i_LIF", "mES_2i_LIF", "mES_FOXA1_ctrl", "mES_POU5F1_ctrl", "mES_SOX2_ctrl", "mES_NT",
                                                "mES_NT",
                                                "mES_NT", "mES_NT", "mES_NT", "mES_NT", "mES_NT", "mES_NT", "mES_NT", "mES_NT",
                                                "mES_NT", "mES_NT", "mES_NT","mES_NT", "mES_NT", "mES_NT", "NPC_DMSO", "NPC_DMSO", "NPC_DMSO"))

dev_scores_dif <- dev_scores %>%
  left_join(condition_match) %>%
  left_join(dev_scores_ctrl) %>%
  mutate(chromVAR_zscore_dif = chromVAR_zscore - ctrl_zscore) %>%
  mutate(chromVAR_zscore_dif_mean = ave(chromVAR_zscore_dif, tf, condition, FUN = function (x) mean(x, na.rm = T)))

# Compute variability
variability <- computeVariability(dev)

plotVariability(variability, use_plotly = FALSE)

saveRDS(dev_scores_dif, "/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/data/chromVAR/dev_scores_mouse_2025_mouse_all.rds")
```

---

## ChromVAR on the human conditions

### Loading data

```{r out.width= "80%", fig.align= "center"}
# Import peak count information for ChromVAR
peak_count <- readRDS("/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/ATAC_seq_analyses/rds/experiment_7784_20241003_human.rds")
assayNames(peak_count) <- "counts"

# Add GC bias
peak_count <- addGCBias(peak_count, genome = BSgenome.Hsapiens.UCSC.hg38)

## Filter peaks - each peak must have at least one count in one condition
peak_count <- filterPeaks(GenomicRanges::sort(peak_count), non_overlapping = TRUE)

motif_ix <- matchMotifs(seq_logos_pwm_all, peak_count, 
                        genome = BSgenome.Hsapiens.UCSC.hg38)
```

---

### Run chromVAR
```{r out.width= "80%", fig.align= "center"}
# Compute deviations
#bg <- getBackgroundPeaks(object = sox2_peak_count_gc_filt)
dev <- computeDeviations(object = peak_count, annotations = motif_ix
                         #,background_peaks = bg
                         )

dev_dif <- differentialDeviations(dev, group = peak_count@colData$treatment)

dev_scores <- deviationScores(dev) %>%
  as.data.frame() %>%
  rownames_to_column(var = "motif_id") %>%
  left_join(motifs_selected) %>%
  pivot_longer(cols = -c(motif_id, tf), names_to = "sample", values_to = "chromVAR_zscore") %>%
  mutate(replicate = gsub(".*(.[1-3]{1}$)", "\\1", sample)) %>%
  mutate(condition = gsub("_.[1-3]{1}$", "", sample)) %>%
  mutate(chromVAR_zscore_mean = ave(chromVAR_zscore, motif_id, condition, FUN = mean)) %>%
  distinct(motif_id, tf, condition, replicate, chromVAR_zscore_mean, chromVAR_zscore)

dev_scores_ctrl <- dev_scores %>%
  filter(condition %in% c("U2OS_DMSO", "MCF7_DMSO")) %>%
  dplyr::select("ref_condition" = condition, "ctrl_zscore" = chromVAR_zscore, tf, replicate) 

condition_match <- data.frame(condition = c("U2OS_PMA", "U2OS_Heat", "U2OS_Calcitriol", "MCF7_Hexestrol"),
                              ref_condition = c("U2OS_DMSO", "U2OS_DMSO", "U2OS_DMSO", "MCF7_DMSO"))

dev_scores_dif <- dev_scores %>%
  left_join(condition_match) %>%
  left_join(dev_scores_ctrl) %>%
  mutate(chromVAR_zscore_dif = chromVAR_zscore - ctrl_zscore) %>%
  mutate(chromVAR_zscore_dif_mean = ave(chromVAR_zscore_dif, tf, condition, FUN = mean))

# Compute variability
variability <- computeVariability(dev)

plotVariability(variability, use_plotly = FALSE)

saveRDS(dev_scores_dif, "/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/data/chromVAR/dev_scores_human.rds")
```






# Session Info
```{r}
paste("Run time: ",format(Sys.time()-StartTime))
getwd()
date()
sessionInfo()
```

