---
title: "Systematic comparison of TF reporter activity with ATAC-seq-inferred TF activity"
author: 
  - name: "Max Trauernicht"
    email: "m.trauernicht@nki.nl"
    affiliation: "Netherlands Cancer Institute - van Steensel lab"
date: '`r format(Sys.time(), "%d/%m/%Y")`'
output: 
  html_document:
    theme: united
    highlight: pygments
    fig_caption: yes
    code_folding: hide
    df_print: kable
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
---

---

### Aim
I performed reporter assays in HepG2 and K562 cells to directly measure the activity of 86 TFs. In parallel, I computed TF activities indirectly from accessibility using ATAC-seq. I now want to systematically compare these two measures in this script. 
The first part of this script describes how the candidate regions for the accessibility analysis were selected. 

---

## Setup {.tabset}

<!-- little HTML script to do indentation of the table of contents -->
<script>
    $(document).ready(function() {
      $items = $('div#TOC li');
      $items.each(function(idx) {
        num_ul = $(this).parentsUntil('#TOC').length;
        $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
      });

    });
</script>

```{css, echo = FALSE}
div.sourceCode {
  overflow-x: hidden;
}
```

### Libraries

```{r setup, out.width= "80%", fig.align= "center", warning= FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
StartTime <-Sys.time()

# 8-digit Date tag:
Date <- substr(gsub("-","",Sys.time()),1,8) 
# libraries:
library(GenomicRanges)
library(rtracklayer)
library(BSgenome.Hsapiens.UCSC.hg38)
library(GenomicFeatures)
library(BSgenome)
library(spgs)
library(dplyr)
library(ggplot2)
library(reshape)
library(ggbeeswarm)
library(biomaRt)
library(readr)
library(stringr)
library(maditr)
library(phylotools)
library(plyr)
library(tidyr)
library(ggpubr)
library(SummarizedExperiment)
```


### Custom functions

```{r, out.width= "80%", fig.align= "center"}
SetFileName <- function(filename, initials) {
  # Set filename with extension and initials to make filename with date integrated.
  filename <- substitute(filename)
  initials <- substitute(initials)
  filename <- paste0(initials, Date, filename)
  filename
}

ReadFasta<-function(file) {
   # Read the file line by line
   fasta<-readLines(file)
   # Identify header lines
   ind<-grep(">", fasta)
   # Identify the sequence lines
   s<-data.frame(ind=ind, from=ind+1, to=c((ind-1)[-1], length(fasta)))
   # Process sequence lines
   seqs<-rep(NA, length(ind))
   for(i in 1:length(ind)) {
      seqs[i]<-paste(fasta[s$from[i]:s$to[i]], collapse="")
   }
   # Create a data frame 
   DF<-data.frame(name=gsub(">", "", fasta[ind]), sequence=seqs)
   # Return the data frame as a result object from the function
   return(DF)
}


# Function to load PWM matrix
get_pwm_feature_matrix <- function(motif_meta_fn, fimo_fn, db = 2) {

  # validate args
  valid_dbs <- 1:2
  if(!db %in% valid_dbs)
    stop('Invalid db (database version). Please use db=1 (maintained for backward compatibility only) or db=2')

  # db=1 is maintained for backward compatibility only
  if(db == 1) {

    # read in motif metadata
    motif_meta    <- read.csv(motif_meta_fn)

    # check whether motif metadata contain essential annotations
    if(!all(c('PWM.ID', 'Cognate.TF') %in% colnames(motif_meta))) {
      message('The motif metadata file does not contain the essential columns PWM.ID and Cognate.TF')
    }

    motif_minimal <- motif_meta[, c('PWM.ID', 'Cognate.TF')]

    # load fimo output --> extract motif id, sequence id and p-value
    df <- read.table(fimo_fn)
    df <- df[, c(1, 2, 7)]

    colnames(df) <- c('PWM.ID', 'seqid', 'pval')

    # add TF id
    df <- merge(df, motif_minimal, by = 'PWM.ID')

    # group motif hits by sequence id
    l <- split(df, df[['seqid']])

    # multiple PWM and multiple hits possible. Reduce hits to one per TF, keeping best p-val only
    l <- lapply(l, function(x) {
      x_by_tf <- split(x, x[['Cognate.TF']], drop = TRUE)
      x_by_tf <- lapply(x_by_tf, function(y) y[which.min(y$pval), ])
      do.call('rbind', x_by_tf)
    })

    # initialize feature matrix
    n_tf          <- motif_minimal[['Cognate.TF']] %>%
      unique %>%
      length
    n_seq         <- length(l)
    pwm           <- matrix(1, nrow = n_seq, ncol = n_tf)
    colnames(pwm) <- (motif_minimal[['Cognate.TF']] %>% unique)

    # replace :: from names of composite motifs
    colnames(pwm) <- str_replace_all(colnames(pwm), '::', '_')

    # fill in feature matrix
    for(i in 1 : n_seq) {
      pwm[i, l[[i]][['Cognate.TF']]] <- l[[i]]$pval
    }

    # -log10 transform
    pwm           <- -1 * log10(pwm)

    # coerce to tib and return
    tib_fimo <- as_data_frame(pwm) %>%
      mutate(id = names(l))
      dplyr::select(id, everything())

  }

  # db = 2 (default)
  else {

    # load metadata
    tib_meta    <- read_csv(motif_meta_fn) %>%
      # extract tf symbol from motif id (Cognate_TF unsafe, it can be empty) and replace :: occurrences
      mutate(tf_symbol = str_remove(ID, '_[0-9]*'),
             tf_symbol = str_replace(tf_symbol, '::', '_')) %>%
      dplyr::select(motif_id = `PWM ID`, tf_symbol)

    # load fimo results
    tib_fimo <- read_tsv(fimo_fn) %>%
      # extract motif id, sequence id and p-value
      dplyr::select(motif_id, sequence_name, pval = `p-value`)

    # add tf symbol to fimo results
    tib_fimo <- tib_fimo %>%
      left_join(tib_meta, by = 'motif_id') %>%
      # remove hits with missing motif id (composite pwms)
      filter(!is.na(tf_symbol))

    # select best hit for each motif and sequence
    tib_fimo <- tib_fimo %>%
      group_by(sequence_name, tf_symbol) %>%
      dplyr::slice(which.min(pval)) %>%
      ungroup()

    # spread into feature matrix
    tib_fimo <- tib_fimo %>%
      mutate(pval = -1 * log10(pval)) %>%
      dplyr::select(-motif_id) %>%
      spread(key = tf_symbol, value = pval, fill = 0, drop = TRUE) %>%
      # perform cosmetics on the id
      mutate(id = sequence_name) %>%
      dplyr::select(-c(sequence_name)) %>%
      dplyr::select(id, everything())

  }

  return(tib_fimo)

}


theme_classic_lines <- function() {
  theme_pubr(border = T, legend = "top") +
            theme(panel.grid.major = element_line(colour = "#adb5bd", size = 0.1),
                  strip.background = element_rect(fill = "#ced4da")
            )
}

theme_classic_lines_45 <- function() {
  theme_pubr(border = T, legend = "top", x.text.angle = 45) +
            theme(panel.grid.major = element_line(colour = "#adb5bd", size = 0.1),
                  strip.background = element_rect(fill = "#ced4da")
            )
}

theme_classic_lines_90 <- function() {
  theme_pubr(border = T, legend = "top", x.text.angle = 90) +
            theme(panel.grid.major = element_line(colour = "#adb5bd", size = 0.1),
                  strip.background = element_rect(fill = "#ced4da")
            )
}

theme_set(theme_classic_lines())

colors_diverse <- c("#264653", "#9AC1AE", "#5D987B", "#f2cc8f", "#e76f51")

ggplot_custom <- function(...) ggplot2::ggplot(...) + 
  scale_color_manual(values = colors_diverse) + 
  scale_fill_manual(values = colors_diverse)
```

## RNA-seq data table prep
```{r}
## Load generated mESC RNA-seq data and calculate TPM
mES_rna <- read_delim("/shared/gcf/m.trauernicht/7813/genecounts.txt") %>%
  filter(gene_biotype == "protein_coding") %>%
  dplyr::select(ensembl_gene_id, "mES_2i_LIF_r3" = `3_mES_2i_LIF_r3`, "mES_2i_LIF_r2" = `2_mES_2i_LIF_r2`,
                "mES_2i_LIF_r1" = `1_mES_2i_LIF_r1`,
                "mES_2i_r1" = `4_mES_2i_r1`, "mES_2i_r2" = `5_mES_2i_r2`,
                "mES_2i_r3" = `6_mES_2i_r3`) %>%
  column_to_rownames("ensembl_gene_id")

write_tsv(mES_rna, "/DATA/usr/m.trauernicht/projects/ATAC_TF_reporter_comparison/RNA_seq_analyses/mES_RNA_counts.tsv")
```



## DiffTF based TF activity computation
```{r}
## DiffTF was run, final output is imported:
difftf_activity <- read_tsv("/DATA/usr/m.trauernicht/projects/SuRE-TF/ATAC_seq_hepg2/diffTF/output_rna_top1000_inverse/FINAL_OUTPUT/extension100/HepG2vsK562.top1000.summary.tsv") %>%
  dplyr::select(TF, 'diffTF_activity' = weighted_meanDifference, 'classification' = classification_q0.001_final)
difftf_activity$classification[difftf_activity$TF == "POU5F1::SOX2"] <- difftf_activity$classification[difftf_activity$TF == "POU5F1"]
difftf_activity$classification[difftf_activity$TF == "RAR:RXR"] <- difftf_activity$classification[difftf_activity$TF == "RARA"]

difftf_activity2 <- read_tsv("/DATA/usr/m.trauernicht/projects/SuRE-TF/ATAC_seq_hepg2/diffTF/output_rna_allTFBS/FINAL_OUTPUT/extension100/HepG2vsK562.top1000.summary.tsv") %>%
  dplyr::select(TF, 'diffTF_activity2' = weighted_meanDifference, 'classification2' = classification_q0.001_final)

difftf_activity3 <- read_tsv("/DATA/usr/m.trauernicht/projects/SuRE-TF/ATAC_seq_hepg2/diffTF/output_rna_exact_match_only/FINAL_OUTPUT/extension100/HepG2vsK562.top1000.summary.tsv") %>%
  dplyr::select(TF, 'diffTF_activity2' = weighted_meanDifference)

# Check if there is a difference between using only direct k-mers as input or using (more degenerate) motif matches
diff <- merge(difftf_activity, difftf_activity3, by = "TF")
ggplot(diff, aes(x = diffTF_activity, y = diffTF_activity2)) +
  geom_point() +
  xlab("DiffTF activity, input = regions similar to motif") +
  ylab("DiffTF activity, input = same sequence as in reporter") +
  geom_abline(linetype = "dashed")



# Import mean reporter activities per TF
reporter_activity_all <- read_csv2("/DATA/usr/m.trauernicht/projects/SuRE-TF/ATAC_seq_hepg2/diffTF/tf_reporter_activities_ernst.csv")
reporter_activity_all_mean <- reporter_activity_all %>%
  unique() %>%
  pivot_wider(names_from = "condition", values_from = c("ernst_activity", "reporter_activity_minP")) %>%
  mutate(dif_ernst = log2(ernst_activity_K562 / ernst_activity_HepG2),
         dif_reporter = log2(reporter_activity_minP_K562 / reporter_activity_minP_HepG2)) %>%
  dplyr::select(dif_ernst, dif_reporter, 'TF' = tf) %>%
  unique()
reporter_activity_all_mean$TF <- gsub("::.*", "_", reporter_activity_all_mean$TF)
reporter_activity_all_mean$TF[reporter_activity_all_mean$TF == "POU5F1_"] <- "POU5F1::SOX2"
reporter_activity_all_mean$TF[reporter_activity_all_mean$TF == "RARA:RXRA"] <- "RAR:RXR"
reporter_activity_all_mean$TF[reporter_activity_all_mean$TF == "ETS1"] <- "ETS2"
reporter_activity_all_mean$TF[reporter_activity_all_mean$TF == "NFATc1"] <- "NFATC1"
reporter_activity_all_mean$TF[reporter_activity_all_mean$TF == "TCF3"] <- "TCF7L2"
reporter_activity_all_mean$TF <- gsub("_", "", reporter_activity_all_mean$TF)


activity_comp <- merge(difftf_activity, reporter_activity_all_mean, by = "TF", all = T) %>%
  na.omit()


# Add RNA-seq data from HepG2 and K562
rnaseq_matrix <- read_tsv("/DATA/scratch/usr/t.v.schaik/proj/3D_nucleus/results/ts210514_RNAseq_Ki67_depletion/results_pe/Star/count_table.tsv") %>%
  dplyr::select('ENSEMBL' = 1, 'HepG2_r1' = 2, 'HepG2_r2' = 8, 'K562_r1' = 7, 'K562_r2' = 13) %>%
  mutate(ENSEMBL = gsub(".[0-9]+$", "", ENSEMBL))

# Add gene names
ensembl <- useMart("ensembl")
mart <- useDataset("hsapiens_gene_ensembl", mart = ensembl)
gene <- rnaseq_matrix$ENSEMBL
G_list <- getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id","hgnc_symbol"),values=gene,mart= mart)
rnaseq_matrix <- merge(rnaseq_matrix,G_list,by.x="ENSEMBL",by.y="ensembl_gene_id")

rnaseq_matrix <- rnaseq_matrix[rnaseq_matrix$hgnc_symbol != "",]

# Filter for genes with no counts
tib_nocounts <- rnaseq_matrix %>%
  dplyr::select(-hgnc_symbol) %>%
  gather(key, value, -ENSEMBL) %>%
  group_by(ENSEMBL) %>%
  dplyr::summarise(count = sum(value > 2) > 1)
idx_nocounts <- which(tib_nocounts$count)

genes <- gene[idx_nocounts]
rnaseq_matrix <- rnaseq_matrix[idx_nocounts, ]

# index <- read.csv("/DATA/usr/m.trauernicht/software/diffTF/src/TF_Gene_TranslationTables/HOCOMOCO_v11/translationTable_hg38.csv", sep = " ")
# rnaseq_matrix <- rnaseq_matrix[rnaseq_matrix$ENSEMBL %in% index$ENSEMBL,]
# rnaseq_matrix <- rnaseq_matrix %>%
#   dplyr::select(-hgnc_symbol)
# write_tsv(rnaseq_matrix, "/DATA/usr/m.trauernicht/data/RNA_seq/hepG2_k562_counts.tsv")

activity_comp <- activity_comp %>%
  mutate(diffTF_activity_norm = (diffTF_activity-mean(diffTF_activity))/sd(diffTF_activity),
         reporter_activity_norm = (dif_reporter-mean(dif_reporter))/sd(dif_reporter))

activity_comp2 <- merge(activity_comp, rnaseq_matrix, by.x = 'TF', by.y = 'hgnc_symbol')
  
# Import class of TF
tf_class <- read_csv2("/DATA/usr/m.trauernicht/projects/SuRE-TF/data/library_design/lambert_2018/garcia_tf-properties.csv") %>%
  dplyr::select(TF, chromatin_regulation_mode)

activity_comp <- merge(activity_comp, tf_class, all = T)
activity_comp$chromatin_regulation_mode[activity_comp$TF == "POU5F1::SOX2"] <- activity_comp$chromatin_regulation_mode[activity_comp$TF == "POU5F1"]
activity_comp$chromatin_regulation_mode[activity_comp$TF == "RAR:RXR"] <- activity_comp$chromatin_regulation_mode[activity_comp$TF == "RARA"]
activity_comp <- activity_comp %>% na.omit()


ggplot(activity_comp, aes(x = reporter_activity_norm, y = diffTF_activity_norm)) +
  geom_point(aes( color = chromatin_regulation_mode)) +
  geom_abline(linetype = "dashed") +
  geom_smooth(method = "lm", color = "grey", fill = "lightgrey") +
  xlab("Differential Reporter Activity (K562/HepG2)") +
  ylab("Differential Accessibility - diffTF (K562/HepG2)") +
  scale_color_manual(values = c("grey",colors_diverse[c(2,5,4)])) +
  facet_wrap(~classification)

for (i in unique(activity_comp$classification)) {
  print(paste("cor =", cor.test(activity_comp$reporter_activity_norm[activity_comp$classification == i], activity_comp$diffTF_activity_norm[activity_comp$classification == i])$estimate,
              "| p-value =", cor.test(activity_comp$reporter_activity_norm[activity_comp$classification == i], activity_comp$diffTF_activity_norm[activity_comp$classification == i])$p.value,
              "| classification =", i))
}


# Which of the TFs are correlating and which not?
activity_comp3 <- activity_comp2 %>%
  mutate(reporter_activity_norm = abs(reporter_activity_norm),
         diffTF_activity_norm = abs(diffTF_activity_norm),
         dif = reporter_activity_norm - diffTF_activity_norm,
         dif_rna = K562_r1 - HepG2_r1) %>%
  dplyr::select(dif, classification, TF, dif_rna) %>%
  unique()

ggplot(activity_comp3 %>% filter(classification != "not-expressed"), 
       aes(x = classification, y = dif)) +
  geom_quasirandom()

activity_comp3 <- activity_comp2 %>%
  mutate(dif_rna = K562_r1 - HepG2_r1) %>%
  dplyr::select(reporter_activity_norm, diffTF_activity_norm, classification, TF, dif_rna) %>%
  unique()

ggplot(activity_comp3 %>% filter(classification != "not-expressed"), 
       aes(x = reporter_activity_norm, y = dif_rna)) +
  geom_point()+
  geom_abline(linetype = "dashed") +
  geom_smooth(method = "lm", color = "grey", fill = "lightgrey") 

ggplot(activity_comp3 %>% filter(classification != "not-expressed"), 
       aes(x = diffTF_activity_norm, y = dif_rna)) +
  geom_point()  +
  geom_abline(linetype = "dashed") +
  geom_smooth(method = "lm", color = "grey", fill = "lightgrey") 



# The same for DMSO vs CDCA
difftf_activity_cdca <- read_tsv("/DATA/usr/m.trauernicht/projects/SuRE-TF/ATAC_seq_hepg2/diffTF/output_fxr/FINAL_OUTPUT/extension100/DMSOvsCDCA.top1000.summary.tsv") %>%
  dplyr::select(TF, 'diffTF_activity' = weighted_meanDifference)

reporter_activity_dif <- reporter_activity %>%
  mutate(reporter_activity = log2(HepG2_CDCA / HepG2_DMSO)) %>%
  dplyr::select(TF, reporter_activity)

activity_comp_cdca <- merge(difftf_activity_cdca, reporter_activity_dif)


ggplot(activity_comp_cdca, aes(x = diffTF_activity, y = reporter_activity)) +
  geom_point(color = ifelse(activity_comp_cdca$reporter_activity > 1, "red", "black")) +
  geom_abline() +
  geom_smooth(method = "lm") +
  xlab("Differential Accessibility - diffTF (CDCA/DMSO)") +
  ylab("Differential Reporter Activity (CDCA/DMSO)") 

```

---

## ChromVAR differential TF activity computation
```{r}
library(chromVAR)
library(motifmatchr)
library(Matrix)
library(SummarizedExperiment)
library(BiocParallel)
set.seed(1252)

```







### Session Info

```{r}
paste("Run time: ",format(Sys.time()-StartTime))
getwd()
date()
sessionInfo()
```

